```{r load_packages_and_df}
# Load the libraries and data used in this Rmd file
library(tidyverse)
library(here)
library(knitr) # for kable
library(tidycensus)
load(here("data", "df.Rda"))
```

```{r variable_ordering}
################################
# Disability Variable Ordering #
################################
base_disability_list <- c("disability_percent_under5", "disability_percent_5to17", "disability_percent_18to34", "disability_percent_35to64", "disability_percent_65to74", "disability_percent_75andup")

# adds a suffix to a string
# used for generating the male and female disability lists
add_suffix <- function(string, suffix) {
  string <- paste0(string, suffix)
  string
}

# create male, female, and unified disability variable ordering list
# and then turn them back into usable vectors
# instead of lists (to avoid the [[]] nonsense and ambiguity with tidyselect)
male_disability_ordering <- base_disability_list %>%
  map(add_suffix, "_male") %>%
  as.character()
female_disability_ordering <- base_disability_list %>%
  map(add_suffix, "_female") %>%
  as.character()
disability_ordering <- c(male_disability_ordering, female_disability_ordering)

###############################
# Education Variable Ordering #
###############################
education_ordering <- c("education_b_percent_highschool_or_less", "education_b_percent_college_or_more",
                        "education_s_percent_less_than_highschool", "education_s_percent_highschool",
                        "education_s_percent_ged", "education_s_percent_some_college", "education_s_percent_college",
                        "education_s_percent_graduate")

############################
# Master Variable Ordering #
############################
misc_variable_ordering <- c("OD_rate", "income_pc_individual", "percent_white", "percent_black", "population", "deaths")
master_variable_ordering <- c(misc_variable_ordering, education_ordering, disability_ordering)
```

```{r grab_county_geometry, echo=FALSE, message=FALSE, warning=FALSE}
base_map <- get_acs("county",
                    state = "OH",
                    table = "B02001",
                    geometry = TRUE,
                    output = "wide")
# reduce down to GEOID and geometry
base_map <- base_map %>%
  select(GEOID, geometry)
```

```{r map_theme}
theme_map <- theme_minimal() +
  theme(axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major = element_blank())
```

# Data

Census data is stored by table, year, geography, and survey. Each survey has a different set of tables, and surveys can differ from year to year [@UnderstandingACSBasics2020]. Surveys are only available for specific geographic regions, with broader, less frequent surveys covering a greater variety of geographies. Geographies include states, counties, census tracts, and school districts [@GeographicAreasCovered2020].

The majority of my independent variables were sourced from the census Annual Community Survey. The Annual Community Survey, or ACS, "provides a detailed portrait of the social, economic, housing, and demographic characteristics of America’s communities" [@UnderstandingACSBasics2020]. ACS data is available on a geographic hierarchy, going from census tracts to the United States as a whole. This thesis uses ACS data aggregated at the county level from the one-year survey [@GeographicAreasCovered2020]. Due to changes in the way that ACS data were collected in 2010, the dataset includes ACS data from the years 2012 to 2018. Tables are not necessarily consistent between years [@UnderstandingACSBasics2020].

The variables "DeathRateCDC" and "UrbanRural" come from CDC data. They are sourced from the "Drug Poisoning Mortality by County" dataset, which is published by the National Center for Health Statistics. The death rate is age-adjusted. The data was paired with the census and overdose data using the FIPS code, which is a GEOID [@calgaryNCHSDrugPoisoning].

## Sample Size

Unfortunately, not all counties in Ohio are included in the Annual Community Survey. Only counties with a population of 65,000 or greater are included in the one year Annual Community Survey [@GeographicAreasCovered2020]. In total, the sample includes 39 out of Ohio's 88 counties. Since the census data is censored based on county population, there is selection bias in the sample. This means that my results may be an inaccurate reflection of reality if there is correlation between a county's population and the county's fatal drug overdose rate. Given that the literature suggests that rural areas have suffered from greater increases in overdose rates, it is likely that there is correlation between county population and the overdose rate, meaning that there is likely selection bias in the dataset.

Counties excluded in the one year Annual Community Survey are available in other U.S. Census Bureau datasets. Counties with at least a population of 20,000 are included in the one year Annual Community Survey supplemental estimates, which provide a more limited version of the tables included in the broader one year Annual Community Survey. Furthermore, the ACS five year estimates include many counties excluded from the one-year estimates, and the decennial survey includes all areas [@UnderstandingACSBasics2020].

```{r included_counties_map}
# commented out because it broke at some point and I did not realize
# uncomment once I get the whole thesis to knit
# df %>%
#   group_by(county, GEOID) %>%
#   summarize(population = mean(population)) %>%
#   full_join(base_map, by = c("GEOID" = "GEOID")) %>%
#   mutate(included = !is.na(population)) %>%
#   ggplot(aes(geometry = geometry, fill = included)) +
#   geom_sf() +
#   labs(fill = "Included in Sample?",
#        title = "Counties included in sample",
#        x = "",
#        y = "") +
#   theme_map +
#   scale_fill_viridis_d()
```