# Literature Review

## Rationality

### Definition of Rationality

While there is significant evidence that consumers are not "rational", the economics discipline is founded upon the theory of consumer rationality[@arcidiaconoConsumerRationalityMultidisciplinary2011]. According to traditional economic theory, individuals making consumption decisions engage in economizing behavior, meaning that they "choose the options that best advance their personal desires and goals at the least possible cost" [@gwartneyMicroeconomicsPrivatePublic2006, p. 11]. This is a form of rational decision-making. Economists abstract away personal desires and goals by viewing them through the lines of utility, which measures the benefit an individual derives from making any particular choice. Utility differs from person to person because each individual has different preferences, making rational decisions inconsistent between individuals. Furthermore, making rational choices does not necessarily mean making the correct choices. People sometimes derive more utility from doing something "wrong", like eating junk food, than doing something "right", like eating sufficient vegetables[@gwartneyMicroeconomicsPrivatePublic2006].

For the theory of consumer rationality to work, consumers must make choices based on a set of consistent preferences that are complete, reflexive, and transitive. Completeness means that given any two goods X and Y, either X is preferred to Y, Y is preferred to X, or X and Y are preferred equally. Reflexivity means that consumers are indifferent between two identical instances of the same good. For example, reflexivity would be violated if a consumer preferred one dime to another. Transitivity means that if a consumer prefers X to Y, and prefers Y to Z, then they must prefer X to Z [@blackConsumerRationality2013].

### Addiction and Rationality

Health economists study addiction using three types of models. Imperfectly rational models of addiction give consumers two incompatible yet internally consistent utility functions. For example, a consumer might have both a farsighted utility function used for making long-term decisions such as buying a house and a short-term utility function for making marginal decisions such as choosing to use heroin to stave off withdrawal [@culyerDictionaryHealthEconomics2005]. This model does a good job of explaining how heroin addicts living on the street can simultaneously express impressive plans for the future while injecting leftover drops of heroin from a dirty cotton using a used needle [@bourgoisRighteousDopefiend2009]. Myopic irrational models have addicts’ utility functions be exclusively shortsighted by using some combination of heavy discounting of future consequences and future consequences being unknown. For example, an addict may not care about long-term health consequences, and even if they do no individual substance user knows what consequences they will face in the future. Finally, there are theories of rational addiction, as exemplified by Becker and Murphy’s work [@culyerDictionaryHealthEconomics2005].

### A Theory of Rational Addiction

While “addictions would seem to be the antithesis of rational behavior”,  Becker and Murphy’s *A Theory of Rational Addiction* attempts to bridge the gap between homo economus and homo sapiens [-@beckerTheoryRationalAddiction1988]. According to the authors, "a good is potentially addictive if increases in past consumption raise current consumption" [@beckerTheoryRationalAddiction1988, 675]. This property is referred to as adjacent complementarity. Adjacent complementarity can also be described as an increase in consumption today increasing the marginal utility of consumption in the future. The level of adjacent complementarity of a good determines the addictiveness of that good [@beckerTheoryRationalAddiction1988]. Given that opioid dependence has been known to occur after only a week of continuous use, and that opioid cravings are so extreme that long-term addicts are unable to "appreciate the intensity of craving when they are not currently experiencing it", opioids are extremely addictive under this definition [@meierPainKillerEmpire2018; @badgerAlteredStatesImpact2007, 685]. This is supported by the medical literature, which finds that heroin is significantly more addictive than any other substance [@nuttDevelopmentRationalScale2007].

Harmful addictions usually feature tolerance and reinforcement. Tolerance means that current consumption is less satisfying when past consumption is greater, while reinforcement means that greater current consumption increases future consumption [@beckerTheoryRationalAddiction1988]. Opioid tolerance makes any given level of consumption less satisfying because it decreases the duration and intensity of euphoria, sedation, and analgesia that opioids induce, which encourages users to increase their consumption over time [@collettOpioidToleranceClinical1998; @morganAnalysisOpioidEfficacy2011]. From a clinical perspective, opioid tolerance occurs through three mechanisms: changes in the body’s metabolism of the drug, changes in the body’s receptors for a drug, and changes in the brain’s response to the stimulus from the drug [@collettOpioidToleranceClinical1998]. The most famous reinforcement mechanism for opioids is withdrawal. Withdrawal is a common term for the symptoms that chronic opioid users feel when they discontinue use [@collettOpioidToleranceClinical1998; @morganAnalysisOpioidEfficacy2011]. These symptoms are deeply unpleasant, and the "fear of withdrawal has been considered to be one of the major forces behind persistent drug abuse in addicts" [@collettOpioidToleranceClinical1998, 62].

Jonathan Gruber and Botond Koszegi expanded on Becker and Murphy's theory of rational addiction in their article "Is Addiction "Rational"? Theory and Evidence".  While Gruber and Koszegi agree with the majority of the traditional theory of rational addiction, they make consumer preferences time-inconsistent. This means that users of addictive substances do not fully factor in the cost of future addiction when making the decision to consume an addictive substance. While time consistency in preferences is an assumption of the rational consumer model, psychological experiments support Gruber and Koszegi's claim that consumers do not have truly time consistent preferences [@kahnemanThinkingFastSlow2011]. Anecdotally, this is reflected in the number of substance users who express a desire to quit while continuing to consume [@bourgoisRighteousDopefiend2009; @gruberAddictionRationalTheory2001]. Under this modified theory of rational addiction, in addition to the externalities of substance use, there is an "internality": users consume more of a harmful substance than they would choose to if the substance did not have addictive properties. These internalities can dwarf the externalities of a substance in size, meaning that corrective policies that only factor in externalities can be significantly smaller than would be optimal[^1] [[@gruberAddictionRationalTheory2001; @gruberSmokingInternalities2002]].

[^1]: Internalities dwarfing externalities is true of cigarettes. Modeling has suggested that pigouvian taxes that include internalities could be multiple times larger than current taxes [@gruberEconomicsTobaccoRegulation2002; @gruberSmokingInternalities2002].

### Policy Implications of Addiction Models

Each model of addiction suggests somehwat different policies. Under the imperfectly rational model, policymakers should try to help addicts and potential future addicts use their far-sighted utility function when considering substance use. In contrast, the myopic model suggests maximizing the price of addictive substances, since under the model addicts do not have a long-view utility function to appeal to, and therefore minimizing acess is the only feasible approach. When using the theory of rational addiction, policymakers should select policies that either prevent initial addiction (e.g. high prices on addictive goods) or help addicted individuals escape the reinforcement trap (e.g. inpatient treatment, detoxes, and MAT) are effective, since the model imagines addicts as rational decision makers that have derive excessive utility from substance use.. While Gruber and Koszegi's modification of the theory of rational addiction does not change the policy implications, it suggests that much larger interventions are needed (e.g. much higher price increases), since the internalities of addiction are often just as large, if not larger, than the externalities, and the original model does not consider internalities.

## History of the Opioid Epidemic

### What are Opioids?

Opioids are a group of powerful analgesic drugs [@kastinHandbookBiologicallyActive2013]. They both alleviate pain and create intense euphoria, and are considered highly addictive [@macyDopesickDealersDoctors2018; @quinonesDreamlandTrueTale2016]. Opioids generate this effects via activating opioid receptors [@kastinHandbookBiologicallyActive2013]. Opium--the original opioid--is processed liquid from opium poppies, and has been used for pleasure, pain alleviation, and controlling dysentry for thousands of years. In the early 1800s, opium was first purified into morphine, which was later transformed into heroin in the 1870s. More opium derivatives, as well as fully synthetic opioids, were developed in the 1900s. Today, the family of opioids includes heroin, morphine, fentanyl, oxycodone, hydrocodone, buprenorphine, and methadone. Unfortunately, these new stronger opioids increased the risk of overdose, which is when a high dose of opioids causes respiratory failure among other things. While opioids are highly addictive and considered dangerous, they remain uniquely effective at treating severe pain [@brummettManagingPainEssentials2013; @rastegarAmericanSocietyAddiction2015].

### America's First Opioid Epidemic

Opium and its derivatives have been consumed in the United States since the nations founding [@trickeyStoryAmerica19thCentury2018]. During the American Civil War, injured soldiers were given morphine to treat their pain [@arabloueiHistoryOpioidsAmerica2019]. At the time, morphine was relatively new, and was considered almost a miraculous panacea. While some physicians noticed and warned others about the dangers of morphine as early as the late 1860s, it was prescribed heavily by doctors until approximately 1900 due to demand from patients and inadequate medical education [@arabloueiHistoryOpioidsAmerica2019; @trickeyStoryAmerica19thCentury2018]. By 1895, approximately 0.5% of Americans were addicted to morphine and other opioids, with addiction being more prevalent among middle and upper class whites, especially white women [@trickeyStoryAmerica19thCentury2018; @whiteSlayingDragonHistory1998]. Eventually, a combination of medical education, changing social norms, and regulation convinced doctors to cut back their opioid prescriptions, ending the epidemic[@ciccaroneFentanylUSHeroin2017; @quinonesDreamlandTrueTale2016; @trickeyStoryAmerica19thCentury2018]. Afterwards, opioid usage became associated with poor people of color--specifically opium with Chinese immigrants and heroin with African Americans--setting the stage for the demonization of opioids that lasted until quite recently [@trickeyStoryAmerica19thCentury2018].

There are many similarities between the current opioid epidemic and the one that ended 120 years ago. Both epidemics were driven by iatrogenic addiction, meaning that users first started with opioids that were legitimately prescribed to them by a physician [@meierPainKillerEmpire2018; @mustoIatrogenicAddictionProblem1985; @quinonesDreamlandTrueTale2016; @trickeyStoryAmerica19thCentury2018]. Both epidemics primarily affected wealthier white Americans [@quinonesDreamlandTrueTale2016; @trickeyStoryAmerica19thCentury2018]. In both cases, doctors were incentivized to acquiesce to demands for opioids from wealthier patients, despite evidence suggesting that it was not a proper or safe treatment [@meierPainKillerEmpire2018; @trickeyStoryAmerica19thCentury2018]. Both epidemics were worsened by new supposedly non-addictive opioids--morphine and heroin in the 19th century and OxyContin in the 21st [@ciccaroneFentanylUSHeroin2017; @quinonesDreamlandTrueTale2016; @trickeyStoryAmerica19thCentury2018]. Finally, in both epidemics addiction was treated as a medical issue, which is in sharp contrast to how opioid addiction was treated during the era in which it primarily affected impoverished people of color [@trickeyStoryAmerica19thCentury2018].

### Changes in Pain Treatment

The current opioid epidemic was ushered in by changing physician attitudes towards pain alleviation. After the first American opioid epidemic, physicians had an extremely negative outlook on opioids. They were rarely prescribed, even to terminally ill patients in hospital [@quinonesDreamlandTrueTale2016]. During this time period, chronic pain was considered out of scope for most physicians. Patients went to specialized multidisciplinary pain centers, where they received a variety of non-opioid treatments before opioids were even considered. If doctors elected to prescribe opioids, they used low strength pills that were combined with less addictive painkillers, such as acetaminophen, which made them difficult to abuse. By the 1980s, the last physicians who had dealt with widespread iatrogrenic addiction had died off, and stigma towards opioids began to fade [@meierPainKillerEmpire2018; @quinonesDreamlandTrueTale2016].

The change in physicians' attitudes towards opioids was accelerated by several factors. Insurance companies seeking to cut costs reduced payments to or stopped reimbursing entirely providers of multidisciplinary pain care due to its relatively high sticker price compared to medication-based treatment. More generally, insurance companies pressured doctors to spend less time with patients, and there is no better way to quickly end a visit than breaking out a prescription pad [@meierPainKillerEmpire2018; @quinonesDreamlandTrueTale2016; @vanzeePromotionMarketingOxyContin2009]. Simultaneously, hospitals began prioritizing patient satisfaction, creating an incentive structure that rewarded handing out opioid prescriptions to anyone who demanded them, including suspected addicts [@quinonesDreamlandTrueTale2016]. Finally, the introduction and marketing of OxyContin further quelled fears of iatrogenic addiction and made opioids prescriptions seem as routine as antibiotics or statins [@meierPainKillerEmpire2018; @quinonesDreamlandTrueTale2016; @vanzeePromotionMarketingOxyContin2009]. By 1995, physician attitudes toward pain had shifted so much that pain became the fifth vital sign, leading to changes in treatment guidelines that encouraged opioid prescriptions for non-cancer chronic pain [@alpertSupplySideDrugPolicy2017; @meierPainKillerEmpire2018; @quinonesDreamlandTrueTale2016].

### The Introduction of OxyContin


```{r}
df %>%
  filter(Year > 2005,
         Year < 2017) %>%
  group_by(County, Year) %>%
  summarize(PopAdjustedRate = mean(`Prescribing Rate`, na.rm = TRUE) * mean(`Population (CDC)`, na.rm = TRUE),
            Deaths = mean(`Overdose Deaths`, na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(Year) %>%
  summarize(PopAdjustedRate = sum(PopAdjustedRate, na.rm = TRUE),
            Deaths = mean(Deaths, na.rm = TRUE)) %>%
  ggplot(aes(x = Year, y = PopAdjustedRate)) +
  geom_point() +
  geom_line() +
  theme_classic() +
  labs(title = "Opioid Prescriptions in Ohio by Year",
       y = "Prescriptions",
       x = "Year")
```

OxyContin is a prescription opioid painkiller brought to market by Purdue Pharmaceuticals in 1996 [@alpertSupplySideDrugPolicy2017; @meierPainKillerEmpire2018; @quinonesDreamlandTrueTale2016]. It is made from oxycodone, a derivative of opium, and can rival morphine in strength [@meierPainKillerEmpire2018; @quinonesDreamlandTrueTale2016]. OxyContin was unlike previous prescription painkillers in several aspects. In contrast to previous medications such as Vicodin, OxyContin was not blended with additional substances such as acetaminophen, which made it much easier to snort and inject. Furthermore, it was available in up to 160 milligram doses, whereas previous painkillers were capped at five to ten milligrams [@quinonesDreamlandTrueTale2016]. Finally, OxyContin featured a time-delay release mechanism, which the FDA allowed Purdue to market as an abuse deterrent, leading to the perception that OxyContin was non-addictive despite a complete lack of clinical studies demonstrating that this was the case [@alpertSupplySideDrugPolicy2017; @eschHowOneSentence2017; @meierPainKillerEmpire2018; @quinonesDreamlandTrueTale2016; @vanzeePromotionMarketingOxyContin2009].

More specifically, OxyContin's original label stated "Delayed absorption as provided by OxyContin tablets, is believed to reduce the abuse liability of a drug" [@eschHowOneSentence2017; @quinonesDreamlandTrueTale2016, 27]. In addition to encouraging more prescriptions from physicians, this label quelled the concerns of pharmacists seeing a huge uptake in opioid prescriptions in the late 1990s. Furthermore, Purdue Pharmaceuticals sponsored continuing medical education for physicians that minimized the risk of opioid addiction. The seminars lauded a supposed study by Porter and Jick, claiming that it demonstrated that less than one percent of patients could become addicted to opioids[^2]. In reality, the "study" was a brief letter that simply provided summary statistics on opioid use from a database that only covered inpatient care, but the gambit was effective at encouraging wanton prescribing of OxyContin [@haneyDoctorWhoWrote; @meierPainKillerEmpire2018; @porterAddictionRarePatients1980; @quinonesDreamlandTrueTale2016; @vanzeePromotionMarketingOxyContin2009; @zhangOneParagraphLetter19802017]. By the turn of the century, however, some activists, physicians, and pharmacists had noticed a major rise in opioid addictions in their communities, and began to warn others [@meierPainKillerEmpire2018].

[^2]: In reality, the only thing that Porter and Jick's letter demonstrated was that opioid dependence can occur after a brief period of use even when used exclusively under medical supervision while in hospital.

Warning signs of the current opioid crisis first arose in Appalachia, where physicians such as Dr. Art Van Zee observed a huge rise in opioid addictions following the introduction of OxyContin [@meierPainKillerEmpire2018; @vanzeePromotionMarketingOxyContin2009; @quinonesDreamlandTrueTale2016]. Communities that had previously only faced alcoholism suddenly had large number of opioid addicts, who were usually much younger--high schools in inpatient rehabilitation was not unheard of. Anecdotally, Dr. Art Van Zee noticed addicts seeking OxyContin by name within one month of it arriving in local pharmacies [@meierPainKillerEmpire2018]. Addicts weren’t the only ones to notice the Purdue’s blockbuster drug—it was the first drug the DEA targeted for monitoring by brand name [@alpertSupplySideDrugPolicy2017]. By the early 2000s, there was an organized coalition based in Appalachia seeking to reformulate or recall OxyContin. The movement failed, however, due to clever legal and political maneuvering by Purdue [@meierPainKillerEmpire2018].

```{r OhioOpioidTimeseriesGraph, fig.cap="Annual fatal drug overdoses in Ohio counties included in sample"}
df %>%
  filter(Year > 1999,
         Year < 2019) %>%
  group_by(County, Year) %>%
  summarize(deaths = mean(`Overdose Deaths`, na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(Year) %>%
  summarize(deaths = sum(deaths, na.rm = TRUE)) %>%
  ggplot(aes(x = Year, y = deaths)) +
    geom_point() +
    geom_line() +
    theme_classic() +
    labs(x = "Year",
         y = "Annual Fatal Drug Overdoses",
         title = "Overdose Deaths by Year",
         subtitle = "Excludes Counties Reporting Fewer than Ten Fatal Overdoses in that Year")
```

Despite the warnings, opioid prescriptions continued to rise throughout the 2000s and early 2010s [@macyDopesickDealersDoctors2018; @meierPainKillerEmpire2018; @quinonesDreamlandTrueTale2016]. In total, there was a 356% increase in prescriptions from 1999 to 2015 [@kruegerWhereHaveAll2017]. In fact, some communities received more prescription painkillers than there were people who could use them, which highlights the high level of diversion of prescription opioids for non-medical use during this period [@alpertSupplySideDrugPolicy2017; @quinonesDreamlandTrueTale2016]. While many doctors treated OxyContin and similar prescription painkillers as routine pharmaceuticals, their impact on patients was more similar to heroin than statins. Approximately one in every 550 patients started on opioid therapy dies from an opioid-related cause, with most deaths occurring within three years of the initial prescription [@friedenReducingRisksRelief2016; @kruegerWhereHaveAll2017]. Furthermore, prescription painkillers often lead to heroin use; approximately 75% of heroin users first misused prescription opioids [@odonnellTrendsDeathsInvolving2017]. OxyContin's devastation would continue unrestrained until its reformulation in the early 2010s [@quinonesDreamlandTrueTale2016].

### OxyContin Reformulation

In 2010, OxyContin was reformulated. The reformulation made it difficult to grind up and snort or inject, which made it less useful for recreational use [@alpertSupplySideDrugPolicy2017; @macyDopesickDealersDoctors2018; @quinonesDreamlandTrueTale2016]. Medical studies find that the formulation decreased recreational Oxycontin abuse between 30 and 50 percent, and decreased overdoses by up to two-thirds [@ciceroEffectAbuseDeterrentFormulation2012; @coplanEffectAbusedeterrentOpioid2016]. This led to the first decrease in prescription opioid and overdose deaths since 1990. Unfortunately, this decrease coincided with an unprecedented rise in heroin overdoses and deaths [@alpertSupplySideDrugPolicy2017]. This is likely because the reformulation made Oxycontin an inferior good compared to heroin, and other changes in prescription painkiller supply made heroin relatively cheaper.

### The Introduction of Black Tar Heroin

Prior to the 1980s, the US heroin market was dominated by white powder heroin trafficked from Asia into the US through New York City. This heroin was heavily adulterated, expensive, and almost exclusively available in impoverished neighborhoods of major metropolitan areas in the Northeast. In the early 1980s, however, black tar heroin started to trickle into Los Angeles from the Mexican state of Nayarit. The "Xalisco Boys"--young men from the town of Xalisco in Nayarit--would migrate to the US with a small quantity of heroin, sell it, and return home. Unlike traditional heroin markets, which operated out of abandoned buildings in major cities, the Xalisco Boys delivered heroin to suburban buyers after arranging purchases through a network of pagers and phones. This led to a demographic shift in heroin consumption, which had been predominantly consumed by poor urban blacks, towards middle and upper class whites. Further differentiating them from traditional heroin distributors, the Xalisco Boys eschewed violence and operated in small, independent cells, which minimized police attention and kept them off the radar for almost two decades. By the 1990s, the Xalisco Boys began to expand to small and medium sized cities without a preexisting heroin market across the United States [@quinonesDreamlandTrueTale2016].

By the 2000s, relatively large heroin markets supplied exclusively with Nayarit black tar existed in cities such as Portland, Columbus, Salt Lake City, Charleston, and Denver, where there had previously been almost no demand. While police in each individual city were aware of the markets, they were not considered a priority, and were not aggressively targeted by the DEA until after the spike in heroin overdose deaths [@quinonesDreamlandTrueTale2016]. While it is difficult to accurately quantify the size of black markets, it is safe to say that heroin sales exploded after the reformulation of OxyContin and the increase in monitoring and regulation of opioid prescriptions [@alpertSupplySideDrugPolicy2017; @macyDopesickDealersDoctors2018; @quinonesDreamlandTrueTale2016]. This is because illegal drug prices increase the further they are sold from distribution centers, and a decrease in prices is almost guaranteed to lead to an increase in sales.

### Regulatory Changes

In response to the opioid epidemic, some states enacted regulations to attempt to reduce the supply of prescription opioids on the gray and black market. For example, states such as Ohio and Kentucky instituted prescription monitoring programs. These programs identified doctor shoppers[^3] as well as physicians prescribing opioids and other medications of concern in very high quantities or to an abnormally high percentage of their patients. Furthermore, some states began targeting pill mills[^4] [@macyDopesickDealersDoctors2018; @popoviciEffectStateLaws2018; @quinonesDreamlandTrueTale2016]. These laws were effective at reducing opioid overdose deaths--both from prescription opioids and heroin--by approximately 2000 a year [@popoviciEffectStateLaws2018]. These laws increased the cost of acquiring prescription painkillers, which limits their use among both already addicted individuals and new users. Unfortunately, the benefits of these laws were limited by their state-specific scope, as some states like Florida became a haven for pill mills and other prescription diversion tactics, which created a steady supply of painkillers for the black market elsewhere [@quinonesDreamlandTrueTale2016]. Furthermore, the introduction of extremely strong--and therefore extremely dangerous--synthetic opioids into the black market opioid supply chain undid these efforts by more than an order of magnitude.

[^3]: Doctor shoppers are individuals who go to multiple doctors seeking specific highly divertable medications

[^4]: Pill mills are essentially physician-owned opioid distributors masquerading as pain management clinics. According to Quinones, dead giveaways include: high fees payable only in cash, large queues in rural areas, and open drug use around the clinic. While these clinics were often shut down, they were quickly replaced until regulations made them difficult to open at all, and the physicians operating them usually faced license suspensions at most unless they were caught working with organized criminal operations or having inappropriate relationships with doctor-shopping patients [@quinonesDreamlandTrueTale2016].

### The Introduction of Fentanyl

Synthetic opioids such as Fentanyl entered the US black market supply chain in the 2010s [@macyDopesickDealersDoctors2018; @quinonesDreamlandTrueTale2016]. These substances are much stronger than conventional opioids like heroin and OxyContin; Fentanyl is 50 to 100 times as powerful as morphine [@cdcFentanyl2019]. Drug dealers and distributors blend synthetic opioids into heroin, or press them into fake pills, because the synthetics are approximately 95% cheaper than heroin [@frankAddressingFentanylThreat2017; @hempsteadSupplySideResponseDeclining2014]. While most opioid addicts dislike the synthetic opioids, they are unable to identify its presence in their drugs without relatively expensive testing kits, which minimizes the risk to drug distributors who chose to cut their product to maximize profit [@carrollExposureFentanylcontaminatedHeroin2017; @sociasEpidemicDeathsFentanyl2017]. When synthetic opioids are introduced to a black market, overdose deaths increase [@hempsteadSupplySideResponseDeclining2014; @odonnellTrendsDeathsInvolving2017; @sociasEpidemicDeathsFentanyl2017].

The introduction of synthetic opioids such as Fentanyl into the black market increases overdose deaths for several reasons. For one, the extreme potency of synthetic opioids means that even tiny amounts of heterogeneity in the final product sold in the street can cause a fatal overdose [@ciccaroneFentanylUSHeroin2017; @hempsteadSupplySideResponseDeclining2014; @marshallEpidemiologyFentanylinvolvedDrug2017; @odonnellTrendsDeathsInvolving2017; @sociasEpidemicDeathsFentanyl2017]. Furthermore, its introduction lowers the price of street opioids, causing both greater consumption by current users and higher numbers of new users [@frankAddressingFentanylThreat2017]. Since fentanyl is a substitue for other opioids, this price decrease increases consumption among all opioid addicts, including those who previously only used prescription painkillers. In addition, non-opioid recreational drugs are sometimes contaminated with Fentanyl and other synthetic opioids, causing fatal opioid overdose in non-opioid users [@fairbairnNaloxoneHeroinPrescription2017; @frankAddressingFentanylThreat2017]. Furthermore, fentanyl overdoses can kill in as little as two minutes, whereas traditional opioids like heroin take 20 to 30 minutes, which dramatically reduces the opportunity window for bystanders and emergency medical services to provide naloxone [@fairbairnNaloxoneHeroinPrescription2017]. Finally, opioids contaminated with Fentanyl dramatically increase the risk of overdose in recreational users, increasing the pool of people liable to overdose from exclusively dependent users [@sociasEpidemicDeathsFentanyl2017].

### Naloxone

Opioid overdoses can be treated with naloxone [@rastegarAmericanSocietyAddiction2015]. Access to naloxone is key to reducing opioid overdose deaths. Community-level naloxone distribution to people who inject drugs began in Chicago in 1996, and has slowly diffused across the country since then. Large decreases in overdose deaths have been observed in cities after the introduction of naloxone; opioid overdose deaths decreased 95% in the 10 years following San Francisco's decision to supply naloxone [@fairbairnNaloxoneHeroinPrescription2017]. Given that naloxone has such a major impact on overdose deaths, access to naloxone is important to understanding the variation in opioid overdose deaths per capita across the United States.

### Medication-Assisted Treatment: The Gold Standard of Opioid Recovery

Medication-assisted treatment, or maintenance therapy, is the gold standard for treating opioid addiction. While opioids are considered dangerous substances, barring overdose, the primary medical concerns of chronic use are all related to how it is used[^5] @rastegarAmericanSocietyAddiction2015]. Since "long-term abstinence after detoxification is the exception rather than the rule", the most effective way to treat individuals with an opioid addiction is to provide them with a long-term maintenance dose of opioids to minimize cravings and withdrawal coupled with counseling services [@rastegarAmericanSocietyAddiction2015, 155]. Maintenance opioids include methadone and buprenorphine, and are much longer lasting than recreational opioids like heroin [@rastegarAmericanSocietyAddiction2015].

[^5]: For example, street users of heroin often suffer from abcesses from improper injections, HIV and other blood-borne diseases from sharing and reusing needles, and systemic infections. None of these conditions are directly caused by opioid use, and all are greatly exacerbated by homelessness and other poor living conditions [@bourgoisRighteousDopefiend2009; @rastegarAmericanSocietyAddiction2015].

### Economy and History of Ohio

```{r mapOhioCounties, fig.cap='Map of Ohio with counties labeled'}
base_map %>%
  mutate(county = str_remove(county, " County, Ohio")) %>%
  ggplot(aes(label = county)) +
  geom_sf() +
  theme_map +
  geom_sf_text(aes(label = county), size = 2.5) +
  labs(title = "Counties in Ohio")
```

Cuyahoga County, which contains Cleveland, was traditionally the only part of Ohio to have an active heroin market, which filtered down to nearby cities such as Toledo [@quinonesDreamlandTrueTale2016]. This helps explain the high rate of heroin usage in the county. While Cleveland was historically a major city--5th in the nation by population at one point--today it exemplifies "deindustrialization, population decline, and entrenched poverty" [@naomilamoreauxDeclineInnovativeRegion2008]. This phenomena is not limited to Cleveland proper. Between 1985 and 2010, manufacturing employment fell by over 50% in Northeast Ohio, which is the region that contains Cuyahoga County and therefore the city of Cleveland [@matthewhrubeyNortheastOhioManufacturing2011]. Broadly speaking, Ohio as a whole has suffered from a decline in unionized manufacturing jobs, which has led to limited opportunities for low education workers in much of the state. Many authors have claimed that the disappearance of high-paying, relatively low-skilled positions led to increased demand for substances such as heroin. For example, in Portsmouth, Ohio, the main employer shut down, and the city now has one of the highest rates of opioid abuse in the country [@quinonesDreamlandTrueTale2016].

## Supply vs Demand Side Explanations

Explanations for the opioid epidemic in the United States can be broadly characterized as either supply-side or demand-side. Supply-siders claim that an increase in the supply of opioids triggered the 21st century epidemic, while demand-siders believe that social and economic factors are responsible. Supply-siders attribute the increase in the overall opioid supply--and therefore opioid deaths--to the rise of prescription painkillers like OxyContin, the simultaneous price decrease and supply increase of heroin outside of major metropolitan areas, and the introduction of synthetic opioids such as Fentanyl into the recreational opioid supply chain. If addicts are myopically irrational, then supply-side factors will play a major role in determining overdose deaths. In contrast, demand-siders focus on changes in pain treatment, poor economic outcomes in particular regions of the United States, and the overall phenomena of deaths of despair. If demand-siders are correct, then economic and demographic indicators should accurately predict fatal opioid overdose rates. On the other hand, if supply-siders are correct, then economic and demographic indicators other than income should not provide much explanatory value, and instead statistics such as the prescribing rate should have the most explanatory power. Under the rational theory of addiction, both supply-side and demand-side explanations are plausible.