# Data and Methods

```{r variableOrdering}
################################
# Disability Variable Ordering #
################################
base_disability_list <- c("disability_percent_under5", "disability_percent_5to17", "disability_percent_18to34", "disability_percent_35to64", "disability_percent_65to74", "disability_percent_75andup")

# adds a suffix to a string
# used for generating the male and female disability lists
add_suffix <- function(string, suffix) {
  string <- paste0(string, suffix)
  string
}

# create male, female, and unified disability variable ordering list
# and then turn them back into usable vectors
# instead of lists (to avoid the [[]] nonsense and ambiguity with tidyselect)
male_disability_ordering <- base_disability_list %>%
  map(add_suffix, "_male") %>%
  as.character()
female_disability_ordering <- base_disability_list %>%
  map(add_suffix, "_female") %>%
  as.character()
disability_ordering <- c(male_disability_ordering, female_disability_ordering)

###############################
# Education Variable Ordering #
###############################
education_ordering <- c("education_b_percent_highschool_or_less", "education_b_percent_college_or_more",
                        "education_s_percent_less_than_highschool", "education_s_percent_highschool",
                        "education_s_percent_ged", "education_s_percent_some_college", "education_s_percent_college",
                        "education_s_percent_graduate")

############################
# Master Variable Ordering #
############################
misc_variable_ordering <- c("OD_rate", "income_pc_individual", "percent_white", "percent_black", "population", "deaths")
master_variable_ordering <- c(misc_variable_ordering, education_ordering, disability_ordering)
```

## Model

In order to assess the determinants of opioid abuse and fatal overdose, this thesis will examine the correlation between county-level economic and social indicators and a county's fatal overdose rate. The fatal overdose rate was selected because quantifying opioid abuse directly is challenging, and there is little data available. In contrast, reliable statistics on the county-level fatal overdose rate are available as far back as 1999 for some counties. While the overdose rate includes overdose deaths from other narcotics, the vast majority of overdose deaths include at least some usage of opioids. Furthermore, it seems safe to assume that there is a strong correlation between the level of opioid abuse in a county and that county's fatal overdose rate. Throughout the rest of this thesis, the fatal overdose rate is modeled as a function of disability rates, education levels, income, racial demographics, opioid prescribing rates, and population density, with year fixed effects, and in some specifications county fixed effects [@hanck10FixedEffects2019].

The modeling function was designed off of the literature review. Many commentators have emphasized that manual labor leads to disability, which leads to prescription painkillers, which leads to opioid addiction and sometimes fatal overdose [@macyDopesickDealersDoctors2018; @meierPainKillerEmpire2018; @quinonesDreamlandTrueTale2016]. Therefore, disability levels and prescribing rates are included in the model. I hypothesize that both will be positively correlated to the fatal overdose rate. Commentators have also claimed that the opioid epidemic has been driven by opioid abuse spreading from a few poor, predominantly black, urban areas to rural America, which is predominantly white [@macyDopesickDealersDoctors2018; @meierPainKillerEmpire2018; @quinonesDreamlandTrueTale2016]. Hence, racial demographics and population density are included in the model. I hypothesize that low population density will be associated with higher fatal overdose rates. The impact of racial demographics is more ambigious, as the literature suggests that areas with a high percentage of people of color likely already had an above-average overdose rate, but that other areas would have greater increases in the overdose rate. Finally, the model includes education levels and income. I hypothesize that education will have a protective effect, but income will be associated with more overdoses, since more income means greater ability to purchase opioids, and therefore more opportunities to overdose.

## Data

The majority of my independent variables were sourced from the census American Community Survey. The American Community Survey, or ACS, "provides a detailed portrait of the social, economic, housing, and demographic characteristics of America’s communities" [@UnderstandingACSBasics2020]. ACS data is available on a geographic hierarchy, going from census tracts to the United States as a whole, with areas with a higher population level appearing in more surveys. This thesis uses ACS data aggregated at the county level [@GeographicAreasCovered2020]. While the one-year survey is preferrable for time series modeling, it only includes counties with a population of 65,000 or greater, which eliminates many Ohio counties [@UnderstandingACSBasics2020; @GeographicAreasCovered2020]. In fact, the one-year survey only includes 39 out of Ohio's 88 counties. Since the counties are selected for inclusion based on county population, there is selection bias in the sample. This means that the results may be an inaccurate reflection of reality if there is correlation between a county's population and the county's fatal drug overdose rate. Given that the literature suggests that rural areas have suffered from greater increases in overdose rates, it is likely that there is correlation between county population and the overdose rate, meaning that there is probably significant selection bias in the one-year survey dataset. Therefore, this thesis will run the same models with both the one-year and five-year surveys.

The fatal overdose rate, prescribing rate, and density level all come from CDC data. They are sourced from the "Drug Poisoning Mortality by County" dataset, which is published by the National Center for Health Statistics. The death rate is available as a crude rate, an age-adjusted rate, and an imputed rate. The model will be tested on all three independent variables. The density level is a qualitative categorical variable ranging from "noncore" to "large central metro". The prescribing rate is indexed to 100. The data was paired with the census and overdose data using the FIPS code, which is a GEOID [@calgaryNCHSDrugPoisoning].

```{r mapCountiesIncludedInSample, fig.cap="Counties included in the data"}
counties_included <- df %>%
  mutate(is_acs1 = if_else(survey == "acs1", TRUE, FALSE),
         is_acs5 = if_else(survey == "acs5", TRUE, FALSE)) %>%
  group_by(County, GEOID) %>%
  summarize(acs1 = sum(is_acs1, na.rm = TRUE),
            acs5 = sum(is_acs5, na.rm = TRUE)) %>%
  filter(!is.na(GEOID)) %>%
  ungroup() %>%
  mutate(included = if_else(acs1 > 0, TRUE, FALSE))

base_map %>%
  left_join(counties_included, by = c("GEOID" = "GEOID")) %>%
  ggplot(aes(geometry = geometry, fill = included)) +
  geom_sf() +
  labs(fill = "Included?",
       title = "Counties Included in ACS1 Sample",
       subtitle = "All counties included in ACS5",
       x = "",
       y = "") +
  theme_map +
  scale_fill_viridis_d()
```

### Variable Summary Statistics

```{r results="asis"}
as.data.frame(df_acs1) %>%
  mutate(`Density: Noncore` = if_else(Density == "Noncore", TRUE, FALSE),
         `Density: Small Metro` = if_else(Density == "Small Metro", TRUE, FALSE),
         `Density: Micropolitan` = if_else(Density == "Micropolitan", TRUE, FALSE),
         `Density: Large Fringe Metro` = if_else(Density == "Large Fringe Metro", TRUE, FALSE),
         `Density: Medium Metro` = if_else(Density == "Medium Metro", TRUE, FALSE),
         `Density: Large Central Metro` = if_else(Density == "Large Central Metro", TRUE, FALSE)) %>%
  select(-Year,
         -GEOID,
         -`Population (CDC)`,
         -`Overdose Deaths`,
         -survey,
         -`Are Overdose Deaths Suppressed?`,
         -`Is the Fatal Overdose Crude Rate Unreliable?`,
         -`Is the Fatal Overdose Crude Rate Suppressed?`,
         -`Is the Fatal Overdose Age-Adjusted Rate Unreliable?`,
         -`Is the Fatal Overdose Age-Adjusted Suppressed?`,
         -`Fatal Overdose Modeled Rate`) %>%
  select(`Fatal Overdose Crude Rate`,
         `Fatal Overdose Age-Adjusted Rate`,
         `Fatal Overdose Imputed Rate` = `Fatal Overdose Rate (Includes Imputed)`,
         `% Black` = `Percent Black`,
         `Density: Noncore`,
         `Density: Small Metro`,
         `Density: Micropolitan`,
         `Density: Large Fringe Metro`,
         `Density: Medium Metro`,
         `Density: Large Central Metro`,
         `Prescribing Rate`,
         `Median Individual Income`,
         `Median Household Income`,
         `% with Highschool or Less` = `Percent with Highschool or Less`,
         `% with Some College` = `Percent with Some College`,
         `% with Bachelors or More` = `Percent with Bachelors or More`,
         `% Disabled: Males Under 5` = `Percent of Males Aged Under 5 with a Disability`,
         `% Disabled: Males 5 to 17` = `Percent of Males Aged 5 to 17 with a Disability`,
         `% Disabled: Males 17 to 34` = `Percent of Males Aged 18 to 34 with a Disability`,
         `% Disabled: Males 35 to 64` = `Percent of Males Aged 35 to 64 with a Disability`,
         `% Disabled: Males 65 to 74` = `Percent of Males Aged 65 to 74 with a Disability`,
         `% Disabled: Males 75 and Up` = `Percent of Males Aged 75 and Up with a Disability`,
         `% Disabled: Females Under 5` = `Percent of Females Aged Under 5 with a Disability`,
         `% Disabled: Females 5 to 17` = `Percent of Females Aged 5 to 17 with a Disability`,
         `% Disabled: Females 17 to 34` = `Percent of Females Aged 18 to 34 with a Disability`,
         `% Disabled: Females 35 to 64` = `Percent of Females Aged 35 to 64 with a Disability`,
         `% Disabled: Females 65 to 74` = `Percent of Females Aged 65 to 74 with a Disability`,
         `% Disabled: Females 75 and Up` = `Percent of Females Aged 75 and Up with a Disability`) %>%
  stargazer(header = FALSE,
            summary.stat = c("mean", "median", "sd", "min", "max"),
            title = "One Year Estimates (218 Observations)",
            font.size = "small",
            digits = 2)
```

```{r results="asis"}
as.data.frame(df_acs5) %>%
  mutate(`Density: Noncore` = if_else(Density == "Noncore", TRUE, FALSE),
         `Density: Small Metro` = if_else(Density == "Small Metro", TRUE, FALSE),
         `Density: Micropolitan` = if_else(Density == "Micropolitan", TRUE, FALSE),
         `Density: Large Fringe Metro` = if_else(Density == "Large Fringe Metro", TRUE, FALSE),
         `Density: Medium Metro` = if_else(Density == "Medium Metro", TRUE, FALSE),
         `Density: Large Central Metro` = if_else(Density == "Large Central Metro", TRUE, FALSE)) %>%
  select(-Year,
         -GEOID,
         -`Population (CDC)`,
         -`Overdose Deaths`,
         -survey,
         -`Are Overdose Deaths Suppressed?`,
         -`Is the Fatal Overdose Crude Rate Unreliable?`,
         -`Is the Fatal Overdose Crude Rate Suppressed?`,
         -`Is the Fatal Overdose Age-Adjusted Rate Unreliable?`,
         -`Is the Fatal Overdose Age-Adjusted Suppressed?`,
         -`Fatal Overdose Modeled Rate`) %>%
  select(`Fatal Overdose Crude Rate`,
         `Fatal Overdose Age-Adjusted Rate`,
         `Fatal Overdose Imputed Rate` = `Fatal Overdose Rate (Includes Imputed)`,
         `% Black` = `Percent Black`,
         `Density: Noncore`,
         `Density: Small Metro`,
         `Density: Micropolitan`,
         `Density: Large Fringe Metro`,
         `Density: Medium Metro`,
         `Density: Large Central Metro`,
         `Prescribing Rate`,
         `Median Individual Income`,
         `Median Household Income`,
         `% with Highschool or Less` = `Percent with Highschool or Less`,
         `% with Some College` = `Percent with Some College`,
         `% with Bachelors or More` = `Percent with Bachelors or More`,
         `% Disabled: Males Under 5` = `Percent of Males Aged Under 5 with a Disability`,
         `% Disabled: Males 5 to 17` = `Percent of Males Aged 5 to 17 with a Disability`,
         `% Disabled: Males 17 to 34` = `Percent of Males Aged 18 to 34 with a Disability`,
         `% Disabled: Males 35 to 64` = `Percent of Males Aged 35 to 64 with a Disability`,
         `% Disabled: Males 65 to 74` = `Percent of Males Aged 65 to 74 with a Disability`,
         `% Disabled: Males 75 and Up` = `Percent of Males Aged 75 and Up with a Disability`,
         `% Disabled: Females Under 5` = `Percent of Females Aged Under 5 with a Disability`,
         `% Disabled: Females 5 to 17` = `Percent of Females Aged 5 to 17 with a Disability`,
         `% Disabled: Females 17 to 34` = `Percent of Females Aged 18 to 34 with a Disability`,
         `% Disabled: Females 35 to 64` = `Percent of Females Aged 35 to 64 with a Disability`,
         `% Disabled: Females 65 to 74` = `Percent of Females Aged 65 to 74 with a Disability`,
         `% Disabled: Females 75 and Up` = `Percent of Females Aged 75 and Up with a Disability`) %>%
  stargazer(header = FALSE,
            summary.stat = c("mean", "median", "sd", "min", "max"),
            title = "Five Year Estimates (257 Observations)",
            font.size = "small",
            digits = 2)
```

We will now examine tables 2.1 and 2.2. The five year estimates include more observations than the one year estimates. The majority of variables are in percentage form in order to isolate the role of county population on the fatal drug overdose rate.. These variables are identified by the "%" in their name. Percentages are calculated dividing the raw count from the Census/CDC by the total population of the county that year. Percentages are stored in the format XX.X%, rather than .XXX%.

The three fatal overdose rate measures have fairly similar values, but the age-adjusted rate is higher than the other two. This makes sense, as it factors how fatal overdose often occurs at ages where other causes of death are uncommon. A typical county in the one year estimates has around 28 people die of an overdose per 100,000. With five year estimates, that number increases to 32. The worst hit counties have approximately three times that death rate, with just over 100 people dying per year per 100,000. The least hit counties have under 10 deaths per year per 100,000. Overall, the overdose death rate is systemically higher with the five year estimates, which suggests that smaller counties have higher overdose rates.

Race is included as the variable percent black. Racial demographic data is represented this way because census demographic data that distinguishes between cacausian and latinx is significantly more challenging to work with, as the categories provided do not neatly add up the total population of the county. Since heroin was considered an urban black drug from the late 1930s until recently, percent black seemed an effective enough variable for assesing the role of race in drug overdose rates [@macyDopesickDealersDoctors2018; @quinonesDreamlandTrueTale2016; @whiteSlayingDragonHistory1998]. The average county is approximately 4% black in the five year estimates, and 8% in the one year estimates. Some counties report zero black residents. The highest percentage of residents that are black is just over 30%.

Density levels vary significantly between the one year and five year estimates. In the one year estimates, no counties are "noncore", whereas 27% are in the five year estimates. This makes sense, as "noncore" counties are very low in population, and therefore are not included in the one year estimates. The rest of this paragraph will use the five year estimates, since they are representative of Ohio as a whole. Excluding "noncore" counties, most counties in Ohio range from medium sized towns to medium sized cities, with over 60% being "micropolitan", "large fringe metro", or "medium metro". The few remaining counties are either small towns or the seats of Cleveland or Columbus.

There is massive variation in the prescribing rate. Prescribing rates are higher and vary less in the one year estimates. In the five year estimates, prescribing rates vary by over 20,000%, which is shocking. With those estimates, a typical county has roughly half the maximum number of prescriptions, and almost hundred times the minimum number of prescriptions.

Income levels do not vary between the one year and five year estimates. This suggests that there is little correlation between income levels and low county population, and that the richest counties appear in both samples. Indidivual income is consistently just over half of household income. The wealthiest counties have over three times the level of income as the poorest counties.

There is significant variation in education levels between counties. While in some counties less than ten percent of the adult population has a college degree, in others more than half of the population does. Education levels are higher in the one year estimates, suggesting that rural, low population counties are less educated on average.

Disability is broken into age and sex bins. These bins are kept small because in order to aggregate them, I would need each county-year pair's population broken down into age and sex bins. The type of disability was ignored, as the literature suggested that painful, physical disabilities would have the greatest impact, and the census-provided disability bins only seperated out deafness and blindness. Unsuprisingly, it appears that disability rates soar after the age of 65. This might be because of individuals gaining access to healthcare through Medicare coverage, and therefore getting a formal diagnosis. Male disability rates are consistently higher than female disability rates until the age of 75. Disability rates are slightly higher in the five year estimates, suggesting that rural, low population counties have higher disability rates than average countues.

```{r overdoseRateDistribution, fig.cap="Distribution of drug overdose rate"}
df %>%
  filter(!is.na(`Fatal Overdose Crude Rate`),
         !is.na(survey)) %>% # drop the NA rows that occur because of the missing 2018 deaths data
  ggplot(aes(x = `Fatal Overdose Crude Rate`)) +
  geom_histogram(bins = 25) +
  theme_minimal() +
  facet_wrap(~survey) +
  labs(y = "",
       x = "Overdose Rate (Deaths per 100,000)",
       title = "Distribution of Overdose Rate") +
  theme(axis.text.y = element_blank(),
        plot.title = element_text(hjust = 0.5))
```

The distribution of the fatal drug overdose rate between counties suggests that censoring may be occuring from the bottom. This is because the right-side tail is significantly longer and fatter than the left-side tail [@hillPrinciplesEconometrics2011]. If it is occuring, however, the censoring is fairly minimal. Overall, the overdose rate is higher with the five year estimates.

```{r raceDistribution, fig.cap="Distribution of percent black by county"}
df %>%
  filter(!is.na(survey)) %>%
  ggplot(aes(x = `Percent Black`)) +
  geom_histogram(bins = 20) +
  theme_minimal() +
  labs(title = "Distribution of Percent Black",
       x = "Percent Black",
       y = "") +
  facet_wrap(~survey) +
  scale_x_continuous(labels = function(x) paste0(x, "%")) + # adds percentage signs; scales::percent assumes .XXX format, while I have XX.X
  theme(axis.text.y = element_blank(),
        plot.title = element_text(hjust = 0.5))
```

The distribution of percent black by county does not indicate any data issues when using one year estimates. While observations do not occur in a perfect Gaussian bell curve, there is no reason to believe that the data is censored. With five year estimates, there is a massive cluster of counties between zero and ten percent black.

```{r bachelorsDegreeeDistribution, fig.cap="Distribution of percent with a bachelors degree or more by county"}
df %>%
  filter(!is.na(survey)) %>%
  ggplot(aes(x = `Percent with Bachelors or More`)) +
  geom_histogram(bins = 20) +
  theme_minimal() +
  labs(title = "Distribution of Percent with a Bachelors Degree or More",
       x = "Percent with Bachelors or More",
       y = "") +
  facet_wrap(~survey) +
  scale_x_continuous(labels = function(x) paste0(x, "%")) + # adds percentage signs; scales::percent assumes .XXX format, while I have XX.X
  theme(axis.text.y = element_blank(),
        plot.title = element_text(hjust = 0.5))
```

There are no apparent data issues with the variable percent with a bachelors degree or more. As seen with the race variable, the five year estimates have a large cluster towards the lower end of the distribution.

```{r mapPercentBlack, fig.cap="Map of Ohio counties by percent black"}
df %>%
  group_by(County, GEOID) %>%
  summarize(`Percent Black` = mean(`Percent Black`, na.rm = TRUE)) %>%
  full_join(base_map, by = c("GEOID" = "GEOID")) %>%
  ggplot(aes(fill = `Percent Black`, geometry = geometry)) +
  geom_sf() +
  theme_map +
  scale_fill_viridis_c() +
  labs(title = "Percent Black by County",
       fill = "% Black")
```

The yellow county is Cuyahoga County, which contains the city of Cleveland. Before the 1990s, Cleveland was considered to be the only region in Ohio to contain a significant heroin market [@quinonesDreamlandTrueTale2016]. Therefore, we would expect Cleveland to have started off with an above average drug overdose death rate. With the rise of the Xalisco boys and the introduction of OxyContin, heroin markets spread across the state [@macyDopesickDealersDoctors2018; @quinonesDreamlandTrueTale2016]. Hence, we would expect Cleveland, and therefore Cuyahoga County, to have a higher than average overdose rate prior to the mid 90s, and for other counties to have a larger increase in their overdose death rate.

## Methods

After determining the broad categories of variables I needed through my literature review, I identified relevant census tables with the help of Mahria Lebow, Reed's data services librarian. We identified tables by searching census table shells for keywords like "disability" and "median income" [@lebowCensusHelp]. After identifying the necessary variables, I used the R package Tidycensus to download census data. The package is an API wrapper for data.census.gov, and allows "R users to return Census and ACS data as tidyverse-ready data frames, and optionally returns a list-column with feature geometry for many geographies". The "list-column with feature geometry" allows the R user to easily draw maps using the downloaded census data and ggplot2 [@walkerTidycensus; @tidycensus2020]. 

Plots, graphs, and maps that appear within this thesis were produced using ggplot2 unless otherwise noted. ggplot2 is a R package that is part of the Tidyverse universe of R packages. The package allows for a R user to quickly produce aesthetically appealing graphs and maps using any tidy dataset. Visualizations made in ggplot2 are designed using a standardized "grammar of graphics", which helps keep the visual appearance of a variety of different types of visualizations consistent [@wickhamGgplot2ElegantGraphics2016]. One major advantage of working with Tidyverse-compliant packages such as ggplot2 and Tidycensus is that they are designed to work together easily. This allows a R user to go directly from pulling down census data to making maps without having to reshape data or otherwise perform tedious data cleaning, wrangling, and preparation tasks [@walkerTidycensus2020; @wickhamGgplot2ElegantGraphics2016].

Initial regressions were performed using ordinary least squares multiple regression. These econometric models are similar to simple linear regression, but allow for multiple explanatory variables. Excluding the intercept, all parameters show the change in the dependent variable given a unit change in that particular explanatory variable *ceterus paribus*, meaning with all other explanatory variables held constant. For this econometric model to work, several assumptions must be fulfilled, such as having homoskedastic errors and no exact collinearity between explanatory variables [@hillPrinciplesEconometrics2011]. While most of the assumptions are fulfilled, it is possible that there is heteroskedacity in the errors, as the census reports higher a margin of error for smaller counties.

In addition to the simple linear regression models, a model with a censored dependent variable is also used. This model, referred to as a Tobit model, is used when there is censoring of the dependent variable [@hareThesisMeetings2019]. Censoring is when a sizable portion of the observations of the dependent variable takes a particular value, often zero [@hareThesisMeetings2019; @hillPrinciplesEconometrics2011]. For example, a national survey of deaths per household would be censored, as most households would record zero deaths for any particular year. When censoring occurs, least squares regression cannot draw an accurate line through the center of the scattered data, leading to biased and inconsistent parameter estimates. To rectify this, Tobit models calculate the likelihood probability by splitting the probabilities between censored and uncensored observations, and then multiplying these probabilties. This accounts for the data falling into two categories--censored and uncensored--and results in less biased and more consistent parameter estimates than ordinary least squares regression models [@hillPrinciplesEconometrics2011]. Censored regression was done using the `censReg` package [@henningsenCensReg2019], In all cases, the regression models were converted into tables using the package Stargazer, which summarizes regressions and prints visually appealing LaTeX tables [@marekhlavacStargazer2018].