# Results

This chapter proceeds with the models described in the previous chapter. The regression tables resulting from the models will be interepted, which leads into policy recommendations. All models include year fixed effects.

## Model Design

```{r regressionFormulaFunction}
regression_formula <- function(rate_variable, income_variable, fixed_effects=NULL) {
  
  # check that rate_variable is valid
  if(!rate_variable %in% c("`Fatal Overdose Crude Rate`",
                          "`Fatal Overdose Age-Adjusted Rate`",
                          "`Fatak Overdose Modeled Rate`",
                          "`Fatal Overdose Rate (Includes Imputed)`")) {
    stop(rate_variable, " is not a valid fatal overdose rate variable. Did you forget to wrap it in ``s like `Fatal Overdose Crude Rate`?")
  }
  
  # check that income_variable is valid
  if(!income_variable %in% c("`Median Individual Income`",
                             "`Median Household Income`")) {
    stop(income_variable, " is not a valid income variable. Did you forget to wrap it in ``s like `Median Individual Income`?")
  }
  
  # excludes 1 from each category (currently 5 and under for both sexes, highschool or less for education)
  formula_base <- paste(
    "`Prescribing Rate`",
    "`Density`",
    "`Percent Black`",
    "`Percent with Some College`",
    "`Percent with Bachelors or More`",
    "`Percent of Males Aged Under 5 with a Disability`",
    "`Percent of Males Aged 5 to 17 with a Disability`",
    "`Percent of Males Aged 18 to 34 with a Disability`",
    "`Percent of Males Aged 35 to 64 with a Disability`",
    "`Percent of Males Aged 65 to 74 with a Disability`",
    "`Percent of Males Aged 75 and Up with a Disability`",
    "`Percent of Females Aged Under 5 with a Disability`",
    "`Percent of Females Aged 5 to 17 with a Disability`",
    "`Percent of Females Aged 18 to 34 with a Disability`",
    "`Percent of Females Aged 35 to 64 with a Disability`",
    "`Percent of Females Aged 65 to 74 with a Disability`",
    "`Percent of Females Aged 75 and Up with a Disability`",
    sep = " + "
  )
  
  # add fixed effects if they're provided
  if(is.null(fixed_effects)) {
    
    regression_formula <- as.formula(paste0(rate_variable, " ~ ", income_variable, " + ", formula_base))
    
  } else {
    
    regression_formula <- as.formula(paste0(rate_variable, " ~ ", income_variable, " + ", formula_base, " + ", fixed_effects))
    
  }
  
  
  # return formula
  regression_formula
  
}
```

```{r produceRegressionFormulas}
# crude rate
crude_rate_formula <- regression_formula("`Fatal Overdose Crude Rate`", "`Median Household Income`", "Year")
crude_rate_acs1 <- lm(crude_rate_formula, df_acs1)
crude_rate_acs5 <- lm(crude_rate_formula, df_acs5)

# age adjusted rate
age_adjusted_rate_formula <- regression_formula("`Fatal Overdose Age-Adjusted Rate`", "`Median Household Income`", "Year")
age_adjusted_rate_acs1 <- lm(age_adjusted_rate_formula, df_acs1)
age_adjusted_rate_acs5 <- lm(age_adjusted_rate_formula, df_acs5)

# with imputed values
imputed_formula <- regression_formula("`Fatal Overdose Rate (Includes Imputed)`", "`Median Household Income`", "Year")
imputed_acs1 <- lm(imputed_formula, df_acs1)
imputed_acs5 <- lm(imputed_formula, df_acs5)

# individual income
crude_rate_formula_individual_income <- regression_formula("`Fatal Overdose Crude Rate`", "`Median Individual Income`", "Year")
crude_rate_individual_acs1 <- lm(crude_rate_formula_individual_income, df_acs1)
crude_rate_individual_acs5 <- lm(crude_rate_formula_individual_income, df_acs5)

# with county Fixed Effects
county_FE_formula <- regression_formula("`Fatal Overdose Crude Rate`", "`Median Household Income`", "Year + County")
county_FE_acs1 <- lm(county_FE_formula, df_acs1)
county_FE_acs5 <- lm(county_FE_formula, df_acs5)
```

### Household vs Individual Income

```{r doHouseholdAndIndividualIncomeDiffer, results="asis"}
# show that household and individual are not significantly different
stargazer(
  crude_rate_acs1,
  crude_rate_individual_acs1,
  crude_rate_acs5,
  crude_rate_individual_acs5,
  covariate.labels = c(
    "Median Household Income",
    "Median Individual Income",
    "Prescribing Rate",
    "Density: Small Metro",
    "Density: Micropolitan",
    "Density: Large Fring Metro",
    "Density: Medium Metro",
    "Density: Large Central Metro",
    "\\% Black",
    "\\% with Some College",
    "\\% with Bachelors or More",
    "\\% Disabled: Males Under 5",
    "\\% Disabled: Males 5 to 17",
    "\\% Disabled: Males 18 to 34",
    "\\% Disabled: Males 35 to 64",
    "\\% Disabled: Males 65 to 74",
    "\\% Disabled: Males 75 and Up",
    "\\% Disabled: Females Under 5",
    "\\% Disabled: Females 5 to 17",
    "\\% Disabled: Females 18 to 34",
    "\\% Disabled: Females 35 to 64",
    "\\% Disabled: Females 65 to 74",
    "\\% Disabled: Females 75 and Up"
  ),
  title = "Does Individual vs Household Income Significantly Change the Model?",
  omit = c("Year", "Constant"),
  column.labels = c("ACS1 Household", "ACS1 Individual", "ACS5 Household", "ACS5 Individual"),
  single.row = TRUE,
  float.env = "sidewaystable",
  sep.width = 0,
  font.size = "small",
  header = FALSE
)
```

Table 3.1 examines how the model changes when individual income is substituted for household income. As will be seen throughout this chapter, changing between the American Community Survey one year and five year estimates has the greatest impact on results. The coefficients for household and individual income point in opposite directions, but given the lack of statistical significance and the incredibly low magnitudes of the coefficients, it almost certainly meaningless. In general, switching between income types has minimal effects on the coefficients within a survey. Moving from household to individual income makes `percent black` significant for one year estimates, but `percent black` is significant with both five year estimate models. The same is true for the `percent with a bachelors degree or more` variable. The `percent of males aged 35 to 64 with a disability` is significant with household and not individual income for one year estimates, but is significant with both income measures when using five year estimates. Due to household income intuitively seeming more relevant than indidividual income and there not being a clear statistical advantage of either variable, future models will use household income.

In all four model specifications, an increase in the opioid prescribing rate is associated with an increase in the fatal opioid overdose rate.

### One Year vs Five Year Estimates

Columns 2 and 4 in table 3.1 examines the differences between modeling with American Community Survey one year and five year estimates. `ACS1` refers to the one year estimates, whereas `ACS5` refers to the five year estimates. There are several significant changes that occur when switching between one and five year estimates. For one, the coefficient for the `prescribing rate` goes down slightly when using five year estimates, although to a degree that is almost unnoticable. In both cases the `prescribing rate` is staistically significant at the 1% level. The significance and impact of the density level dummies changes as well. With one year estimates, medium density areas are associated with a significantly higher overdose rate. In contrast, density has a statistically significant protective effect when using five year estimates. This is likely due to the five year estimates containing more small, rural counties, which the literature suggested have the highest overdose rates in Ohio. The coefficient of "Percent Black" more than doubles when the five year estimates are used, and is only significant with that dataset. While the coefficient for `percent with a bachelors degree or more` is close to zero and statistically insignificant with one year estimates, it has a statistically significant protective effect when using five year estimates. With five year estimates, as the `percent of males aged 18 to 34 with a disability` rises, the fatal opioid overdose rate rises. The protective effect of the `percent of males aged 36 to 64 with a disability` increases in both magnitude and statistical significance when using five year estimates instead of one year estimates. With five year estimates, the `percent of males aged 65 to 74 with a disability` is associated with increased overdose deaths. Switching from one year to five year estimates increases the magnitude and statistical significance of the protective effect of disabilities among females aged under five and five to 17. Furthermore, using five year estimates increases the coefficient and significance of the `percent of females aged 35 to 64 with a disability` variable. Finally, the five year estimates have greater explanatory power than the one year estimates. I will emphasizing the five year estimates throughout the rest of my thesis because the models using the five year estimates have greater R-squared values than the models using one year estimates, meaning that the models using five year estimates explain more of the variance in the dataset.

### Fatal Overdose Rate Metrics

```{r doCrudeAgeImputedDifferACS5, results="asis"}
stargazer(
  crude_rate_acs5,
  age_adjusted_rate_acs5,
  imputed_acs5,
  covariate.labels = c(
    "Median Household Income",
    "Prescribing Rate",
    "Density: Small Metro",
    "Density: Micropolitan",
    "Density: Large Fring Metro",
    "Density: Medium Metro",
    "Density: Large Central Metro",
    "\\% Black",
    "\\% with Some College",
    "\\% with Bachelors or More",
    "\\% Disabled: Males Under 5",
    "\\% Disabled: Males 5 to 17",
    "\\% Disabled: Males 18 to 34",
    "\\% Disabled: Males 35 to 64",
    "\\% Disabled: Males 65 to 74",
    "\\% Disabled: Males 75 and Up",
    "\\% Disabled: Females Under 5",
    "\\% Disabled: Females 5 to 17",
    "\\% Disabled: Females 18 to 34",
    "\\% Disabled: Females 35 to 64",
    "\\% Disabled: Females 65 to 74",
    "\\% Disabled: Females 75 and Up"
  ),
  title = "How does the type of overdose rate metric influence the results (AS5)?",
  omit = c("Year", "Constant"),
  column.labels = c("Crude", "Age-Adjusted", "Imputed"),
  single.row = TRUE,
  float.env = "sidewaystable",
  sep.width = 0.5,
  font.size = "small",
  header = FALSE
)
```

Table 3.3 shows the differences between using the crude fatal overdose rate, the age-adjusted fatal overdose rate, and the imputed overdose rate when working with American Community Survey five year estimates. While there are minor differences between these model specifications, they all return very similar results. The crude and imputed models include approximately 100 more observations than the age-adjusted model, which makes it difficult to separate out the influence of using an age-adjusted rate from the influence of decreased observations. Adjusting the fatal overdose rate for age decreases the signficiance of the prescribing rate to the 10% level from the 1% level. With all three models, the prescribing rate has a small, but significant, protective effect. Using the age-adjusted fatal overdose rate increases the protective effect of population density on fatal overdoses, but decreases the significance of that effect. This probably means that there is correlation between the age distribution of a county and its population density. Using the age-adjusted rate makes all male disability rates insignificant, and increases the impact of female disability rates. This may be correlated with the disability rates being broken down into age bins. The imputed rate produces results are almost identical to the crude rate results. This is unsuprising, as the imputed rate is simply the crude rate with a few more observations. Given that using the age-adjusted rate reduces the number of observations significantly, and the imputed rate only provides two more observations while significantly complicating the dependent variable, the rest of this thesis will use the crude rate.

### County Fixed Effects

```{r doCountyFixedEffectsMatter, results="asis"}
stargazer(
  crude_rate_acs1,
  county_FE_acs1,
  crude_rate_acs5,
  county_FE_acs5,
  covariate.labels = c(
    "Median Household Income",
    "Prescribing Rate",
    "Density: Small Metro",
    "Density: Micropolitan",
    "Density: Large Fring Metro",
    "Density: Medium Metro",
    "Density: Large Central Metro",
    "\\% Black",
    "\\% with Some College",
    "\\% with Bachelors or More",
    "\\% Disabled: Males Under 5",
    "\\% Disabled: Males 5 to 17",
    "\\% Disabled: Males 18 to 34",
    "\\% Disabled: Males 35 to 64",
    "\\% Disabled: Males 65 to 74",
    "\\% Disabled: Males 75 and Up",
    "\\% Disabled: Females Under 5",
    "\\% Disabled: Females 5 to 17",
    "\\% Disabled: Females 18 to 34",
    "\\% Disabled: Females 35 to 64",
    "\\% Disabled: Females 65 to 74",
    "\\% Disabled: Females 75 and Up"
  ),
  title = "Final Models",
  omit = c("Year", "County", "Constant"),
  column.labels = c("ACS1 without FE", "ACS1 with FE", "ACS5 without FE", "ACS5 with FE"),
  single.row = TRUE,
  float.env = "sidewaystable",
  sep.width = 0,
  font.size = "small",
  header = FALSE
)
```

Among other things, table 3.5 examines what changes occur in the model when county fixed effects are included. Including county fixed effects increases the R-squareds by a little over 50%. This is unsurprising, as adding cross-sectional unit fixed effects typically causes a significant increase in the R-squared. As before, the five year estimates explain more of the variance in the sample than the one year estimates. With five year estimates and county fixed effects, the model is able to explain over 80% of the variation in the fatal overdose rate. Including fixed effects changes the coefficients to the point of reversing the signs in some cases. These changes will be examined in the following paragraphs.

We will now extract our final results from table 3.5. Regardless of which specification is used, median household income has no effect on the fatal overdose rate. Before adding county fixed effects, an increase in the prescribing rate is associated with an increase in the fatal overdose rate. This is true of both one year and five year estimates. After controlling for fixed effects, an increase in the prescribing rate is correlated with a decrease in the fatal overdose rate. These relationships are statistically significant at the one percent level for all models other than county fixed effects with five year etimates. The change in the sign of the coefficients implies that some unobservable variable is driving both the prescibing rate and the overdose rate, and that there is not a direct causal relationship between prescription opioids and overdoses. While the literature emphasizes that there is a relationship, the dataset used in this thesis starts around the time that the pill-mill crisis in Ohio started to subside, which could potentially explain the disconnect.

Without county fixed effects, an increase in the percentage of the population that is black is associated with an increase in the fatal overdose rate. This is true of both one year and five year estimates, although the effect is stronger and statistically significant solely when using the five year estimates. After including the county fixed effects, the coefficients are reversed. Neither one year or five year estimates are statistically significant, and the five year estimate coefficient is approximately ten times as large as the one year estimate coefficient. This suggests that after controlling for unobservables, counties that have an increase in the percentage of their population that is black have decreased overdose deaths. It is possible that Cuyahoga county, which has both an extremely high rate of opioid overdoses and one of the largest percentage of people of color of any Ohio county, causes this reversal. There is significant evidence that physicians discriminated against people of color during the pill mill boom, which led to a lower rate of opioid prescriptions among people of color, which ultimately led to a lower rate of opioid abuse and therefore overdose, although said racial discrimination likely caused a higher rate of misuse among people of color who did receive prescription painkillers [@fraktRareCaseWhere2019; @swiftRacialDiscriminationMedical2019].

Like many other variables, including county fixed effects dramatically changes the coefficients of the education variables. Without fixed effects, both having some level of college education and having a bachelors degree has a protective effect against fatal overdoses, although the only statistically significant effect is that of having a bachelors or more when using five year estimates. After including county fixed effects, an increase in education levels is associated with an increase in fatal overdoses. This is only statistically significant when looking at `some college` with one year estimates. In all cases, the coefficients are of a rather small magnitude. The change in the sign of the coefficients when county fixed effects are included means that from a cross section perspective, education has a protective effect against fatal overdose, but from a per-county time series perspective, education increases the rate of fatal overdose.

This paragraph will examine the effect of male[^1] disability rates. For ages five to seventeen, an increase in the rate is correlated with an increase in the fatal overdose rate. This effect is stronger when county fixed effects are included, which suggests that fluctuations in the rate within counties matters more than differences in the rate between counties. The effect is only statistically significant with fixed effects and one year estimates. The correlation is also true for ages 18 to 34, although the within versus between distinction is more ambigious, and the results are also statistically significant with five year estimates without county fixed effects. For males aged 35 to 64, the disability rate has a protective effect which is statistically significant in three out of the four specifications. This contradicts the prevailing media narrative, which claims that a concentration of manual labor jobs in some areas led to an increased rate of disability among the older, predominantly male people who worked those jobs in those areas, which ultimately led to those areas having extremely high rates of opioid abuse and overdose. For those between the ages of 65 and 74, there is no meaningful effect with one year estimates. With five year estimates, an increase in the disability rate leads to an increase in fatal overdoes in the cross section dimension, and a decrease in the time series dimension. Both effects are statistically significant at the five percent level.

[^1]: This section uses the terms "male" and "female" to indicate that the dataset only captures biological sex at birth, and not gender identity.

We will know look at female disability rates. For those under the age of five, an increase in the disability rate is correlated with an increase in the fatal overdose rate for all models except for five year estimates coupled with county fixed effects. That model is the only statistically significant one, however. For those aged 5 to 17, the disability rate has a protective effect between counties but is associated with an increase in fatal overdoses within counties. The protective effect between counties potentially reflects some underlying difference between how disabilities are diagnosed in different counties. For females aged 18 to 34, disabilities appear to have a protective effect against fatal overdose. The effect is statistically significant and larger in magnitude when county fixed effects are not included. In contrast, an increase in the disability rate for females aged 35 to 64 appears to increase the rate of fatal overdose. This is statistically significant in three out of the four specifications. Overall, this suggests that a diagnosed disability has a protective effect for younger females but a harmful effect for older females. I suspect that the protective effect among younger females reflects increased medical capacity and competence rather than being directly causal, especially since the effect is much larger between counties than within counties. In contrast, the coefficients for older females do not unambigiously change when county fixed effects are included, which means that the effect matters both between and within counties. This suggests that there is a direct causal relationship between the disability rate and the overdose rate. Curiously, neither the media nor the literature focuses on female disability rates, despite these results being less ambigious than the male disability rates.

Finally, we will look at density. The models using the one year estimates do not have coefficients for small metro because the one year estimates do not include any counties with a population under 65,000. With the five year estimates, that density level is associated with lower overdose deaths both with and without county fixed effects. Micropolitan also has a protective effect which gets dramatically stronger with five year estimates and county fixed effects. This suggests that changes in density within counties matter much more than between counties. The same is true for large fring metro, although the one year estimates without county fixed effects show a statistically significant increase in overdose deaths. Excluding column 1, medium metros also have a protective effect which gets dramatically stronger when county fixed effects are included. This is statistically significant for all models. Finally, large central metros appear to have higher than average overdose deaths, especially after including county fixed effects. In general, it seems like there is a barbell effect, where the extremes of density--very rural and very urban--are correlated with higher overdose rates than other density levels.

### Censored Regression

In order to address potential data censoring, I also tried running a Tobit model, which is a form of censored regression. I was unable to get tobit estimates, however, due to the large number of independent variables relative to the number of observations, as well as ambigious censoring[^2]. Regressions ran with one year estimates, fewer variables, and neither county nor year fixed effects did converge, however, and showed no signficiant differences between an OLS and censored regression model. This indicates that there is little to no censoring in the data.

[^2]: the `censReg` package returns "Error: there are no censored observations" unless I engage in data-hacking

```{r tobit, eval=FALSE}
# disabled because no censored regressions
crude_rate_household <- regression_formula("`Fatal Overdose Age-Adjusted Rate`", "`Median Household Income`")
crude_rate_household_FE <- regression_formula("`Fatal Overdose Age-Adjusted Rate`", "`Median Household Income`", "Year")
# picked left basically at random to get rid of "there are no censored observations" error
stargazer(censReg(crude_rate_household, data = df_acs1, logLikOnly = FALSE),
          censReg(crude_rate_household, data = df_acs5, logLikOnly = FALSE),
          censReg(crude_rate_household_FE, data = df_acs1, logLikOnly = FALSE),
          censReg(crude_rate_household_FE, data = df_acs5, logLikOnly = FALSE),
          covariate.labels = c(
    "Median Household Income",
    "Prescribing Rate",
    "Density: Small Metro",
    "Density: Micropolitan",
    "Density: Large Fring Metro",
    "Density: Medium Metro",
    "Density: Large Central Metro",
    "\\% Black",
    "\\% with Some College",
    "\\% with Bachelors or More",
    "\\% Disabled: Males Under 5",
    "\\% Disabled: Males 5 to 17",
    "\\% Disabled: Males 18 to 34",
    "\\% Disabled: Males 35 to 64",
    "\\% Disabled: Males 65 to 74",
    "\\% Disabled: Males 75 and Up",
    "\\% Disabled: Females Under 5",
    "\\% Disabled: Females 5 to 17",
    "\\% Disabled: Females 18 to 34",
    "\\% Disabled: Females 35 to 64",
    "\\% Disabled: Females 65 to 74",
    "\\% Disabled: Females 75 and Up"
  ),
          title = "Tobit: Age-Adjusted Rate on Household Income with and without Year Fixed Effects",
          dep.var.caption = "Fatal Overdose Rate (Age-Adjusted)",
          omit = "Year",
          column.labels = c("ACS1", "ACS5", "ACS1 with FE", "ACS5 with FE"),
          add.lines = list(c("Fixed Effects?", "No", "No", "Yes", "Yes")),
          single.row = TRUE,
          sep.width = 0,
          font.size = "small",
          float.env = "sidewaystable",
          header = FALSE)
```

## Policy Implications

While public policy around the opioid epidemic has emphasized reducing the number of opioid prescriptions, my results indicate that prescribing rates are no longer relevant within the state of Ohio. This suggests that relaxing prescribing guidelines to increase access to effective pain treatment might be reasonable, but it would be prudent to wait for more evidence before doing so, especially given the extensive clinical research showing the addictiveness of opioids. Likewise, the coefficients for education are ambigious, so the usual policy prescription of increasing education to combat any woe is not relevant.

The results suggest that decreasing the number of people who live in extremely high or low density areas would decrease the overdose rate. This goldilocks effect would be difficult to achieve outside of the long-run however, as housing stock changes slowly. In the short-run, further research could potentially identify what about medium density is so benefecial (or what about extremely dense and extremely rural is so harmful), and try to bring those elements to more communities in Ohio. Previous research into social capital and the influence of community on health will be useful for guiding this future research.

While many of the disability rate variables returned ambigious coefficients, several were unambigious enough to create clear policy prescriptions out of. While the media has emphasized the role of male disability from blue-collar employment on increased opioid abuse in the past two decades, the regressions clearly indicate that male disability has either no statistical role in overdose deaths or protects against them. Therefore, policymakers should put less of their effort into male disability and instead focus on female disability. The results show that for females, the disability rate among those aged 18 to 34 has a protective effect, whereas the rate among those aged 35 to 64 increases overdose deaths. Given the harms associated with the older female disability rate, resources should be redirected towards members of that population in order to curtail deaths. Programs could include targeted opioid abuse education and increased access to holistic pain management among other services. Furthermore, researchers should investigate why the disability rate among females aged 18 to 34 has a protective effect. I suspect it is due to underlying county-level factors that make early diagnoses more common. If further research indicates that is the case, then those factors should be spread to other counties. I would hazard a guess that access to quality healthcare with minimal opportunity cost is key.