`r if(knitr:::is_latex_output()) '\\appendix'`

`r if(!knitr:::is_latex_output()) '# (APPENDIX) Appendix {-}'` 

<!--
If you feel it necessary to include an appendix, it goes here.
-->


# Code

```{r ref.label='include_packages', results='hide', echo = TRUE}
```

```{r ref.label=knitr::all_labels(), echo = T, eval = F}
# copied from https://yihui.org/en/2018/09/code-appendix/
```

**In Chapter \@ref(ref-labels):**

```{r ref.label='include_packages_2', results='hide', echo = TRUE}
```

# Timeline

1980s: Last physicians from first opioid epidemic retire

1995: Pain becomes fifth vital sign in physician guidelines

1996: Oxycontin is brought to market by Purdue Pharmaceuticals

1999: VA adopts pain as fifth vital sign [@jonesBriefHistoryOpioid2018]

2001: First major book on prescription painkiller abuse is written (Meier's *Painkiller*)

2000s: Heroin from Nayarit available in most cities

2007: Purdue (the Oxycontin manufacturers) pleads guilty to misbranding the drug [@jonesBriefHistoryOpioid2018; @quinonesDreamlandTrueTale2016]

2010: Oxycontin is reformulated

2010s: Fentanyl and other synthetic opioids enter the black market supply chain

2016: FDA begins to require CME for physicians prescribing opioids [@jonesBriefHistoryOpioid2018]

2017: US government labels the epidemic as a public health emergency[@jonesBriefHistoryOpioid2018]

2019: HEALing Communities Study announced by the National Institute of Health