---
author: 'Jackson Maillie Luckey'
date: 'May 2020'
institution: 'Reed College'
division: 'History and Social Sciences'
advisor: 'Denise Hare'
department: 'Economics'
degree: 'Bachelor of Arts'
title: 'Ohio Opioid Thesis'
knit: "bookdown::render_book"
site: bookdown::bookdown_site
output: 
  thesisdown::thesis_pdf: default
#  thesisdown::thesis_gitbook: default
#  thesisdown::thesis_word: default
#  thesisdown::thesis_epub: default
# If you are creating a PDF you'll need to write your preliminary content (e.g., abstract, acknowledgements) here or
# use code similar to line 22-23 for the .RMD files. If you are NOT producing a PDF, you can delete or silence lines 21-32 in this YAML header.
abstract: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-abstract.Rmd"), collapse = '\n  ')`
# If you'd rather include the preliminary content in files instead of inline
# like below, use a command like that for the abstract above.  Note that a tab is 
# needed on the line after the `|`.
acknowledgements: |
  I want to thank a few people.
dedication: |
  You can have a dedication here if you wish. 
preface: |
  This is an example of a thesis setup to use the reed thesis document class
  (for LaTeX) and the R bookdown package, in general.
# Specify the location of the bibliography below
bibliography: bib/thesis.bib
# Download your specific csl file and refer to it in the line below.
csl: csl/chicago-fullnote-bibliography.csl
lot: true
lof: true
# If you prefer blank lines between paragraphs, un-silence lines  40-41 (this requires package tikz)
#header-includes:
#- \usepackage{tikz}
---

```{r include_packages, include = FALSE}
# This chunk ensures that the thesisdown package is
# installed and loaded. This thesisdown package includes
# the template files for the thesis.
if(!require(devtools))
  install.packages("devtools", repos = "http://cran.rstudio.com")
if(!require(thesisdown))
  devtools::install_github("ismayc/thesisdown")
library(thesisdown)
```

# Introduction {.unnumbered}

The current pioid epidemic is the worst drug epidemic in US history. While opioids are a legitimate medical treatment for acute pain, they are highly addictive and come with the risk of abuse, dependence, and fatal overdose [@aliprantisOpioidsLaborMarket2018]. Abuse of and dependence on opioids has serious consequences. Since 2016, drug overdoses have been the leading cause of accidental death in the United States [@maguireOpioidCrisisAppalachia2019]. Overdoses kill more people than gun violence, motor vehicle incidents, or the HIV/AIDS epidemic did at its peak [@quinonesDreamlandTrueTale2016; @ciccaroneFentanylUSHeroin2017; @alpertSupplySideDrugPolicy2017]. In fact, the scourge of opioid overdoses is so severe that it has lead to an increase in all-cause mortality for white Americans despite mortality decreasing for all other Americans and citizens of comparably developed countries [@caseMortalityMorbidity21st2017]. Reducing the number of deaths due to opioids is essential to reversing this trend, increasing the labor supply, and improving outcomes for millions of Americans.

```{r grab_county_geometry, echo=FALSE, message=FALSE, warning=FALSE}
base_map <- get_acs("county",
                    state = "OH",
                    table = "B02001",
                    geometry = TRUE,
                    output = "wide")
# reduce down to GEOID and geometry
base_map <- base_map %>%
  select(GEOID, geometry)
```

```{r map_theme}
theme_map <- theme_minimal() +
  theme(axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major = element_blank())
```