---
author: 'Jackson Maillie Luckey'
date: 'May 2020'
institution: 'Reed College'
division: 'History and Social Sciences'
advisor: 'Denise Hare'
department: 'Economics'
degree: 'Bachelor of Arts'
title: 'Social and Economic Determinants of the Fatal Overdose Rate in Ohio'
knit: "bookdown::render_book"
site: bookdown::bookdown_site
output: 
  thesisdown::thesis_pdf: default
abstract: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-abstract.Rmd"), collapse = '\n  ')`
acknowledgements: |
  First of all, I would like to thank my thesis adviser, Denise Hare, for supporting me throughout the thesis process. Her advising has carried my though the thesis process, and I would not gotten nearly as far without her. I would also like to thank Nick Wilson for being my first reader, giving me invaluable feedback throughout the thesis project, and for teaching me health economics.

  Reed staff have been just as important as faculty in supporting my thesis. In particular, I would like to thank Kristin Bott and Mahria Lebow. I learned the R skills that provided the backbone of this thesis primarily through working for Kristin in Reed's mLab. Furthermore, she has connected me with resources at Reed as I ran into issues working through my thesis. Reed's data services librarian, Mahria Lebow, helped me navigate census data and saved me countless hours of futile struggling with the census API. Without their help, I would likely still be trying to gather my data.

  I would like to thank my father, Bill Luckey, for introducing me to public health policy and the study of addiction. Without his influence in my life, I would not have gone down this path. I'd also like to thank my mom for helping me get through my years at Reed, and giving me a safe place to stay during the 2019-2020 coronavirus pandemic. Without her, I never would have finished my thesis--or my degree.

  I would also like to thank Arnie Aldridge, Brendan Wedehase, and Gary Zarkin for their invaluable role in my thesis. Without their guidance during my internship at RTI, I would have never developed the technical skills nor background knowledge that was necessary for producing this document.
  
  Finally, I would like to thank everyone at FEI for introducing me to the current state of substance use disorder policy and treatment research.
preface: |
  `r if(knitr:::is_latex_output()) paste(readLines("00--prelim.Rmd"), collapse = '\n  ')`
# Specify the location of the bibliography below
bibliography: bib/thesis.bib
# Download your specific csl file and refer to it in the line below.
csl: csl/chicago-note-bib.csl
lot: true
lof: true
header-includes:
  \usepackage{tikz}
---

```{r includePackages, include = FALSE}
# This chunk ensures that the thesisdown package is
# installed and loaded. This thesisdown package includes
# the template files for the thesis.
if(!require(devtools))
  install.packages("devtools", repos = "http://cran.rstudio.com")
if(!require(thesisdown))
  devtools::install_github("ismayc/thesisdown")
library(thesisdown)
```

```{r loadPackages, include=FALSE}
# Load the libraries and data used in this Rmd file
library(tidyverse)
library(broom) # for manipulating regression output
library(here)
library(knitr) # for kable
library(kableExtra)
library(forcats)
library(citr)
library(rvest)
library(glue)
library(censReg)
library(stargazer) # for regression output
library(tidycensus) # putting this last stops the intermittent st_drivers() error
options(knitr.table.format.args = list(scientific = FALSE,
        big.mark = ",",
        digits = 5),
        knitr.table.booktabs = T)
opts_chunk$set(include = TRUE,
               echo = FALSE,
               message = FALSE,
               warning = FALSE) # have R code only appear in appendix
```

# Introduction {.unnumbered}

The current opioid epidemic is the worst drug epidemic in US history. While opioids are a legitimate medical treatment for acute pain, they are highly addictive and come with the risk of abuse, dependence, and fatal overdose [@aliprantisOpioidsLaborMarket2018]. Abuse of and dependence on opioids has serious consequences. Since 2016, drug overdoses have been the leading cause of accidental death in the United States [@maguireOpioidCrisisAppalachia2019]. Overdoses kill more people than gun violence, motor vehicle incidents, or the HIV/AIDS epidemic did at its peak [@quinonesDreamlandTrueTale2016; @ciccaroneFentanylUSHeroin2017; @alpertSupplySideDrugPolicy2017]. In fact, the scourge of opioid overdoses is so severe that it has led to an increase in all-cause mortality for white Americans despite mortality decreasing for all other Americans and citizens of comparably developed countries [@caseMortalityMorbidity21st2017]. Reducing the number of deaths due to opioids is essential to reversing this trend, increasing the labor supply, and improving outcomes for millions of Americans.

In this thesis, I will start by reviewing the literature. The literature review will open by defining rationality in the context of economics. I will follow this up by examining the different models of addiction and how they approach rationality, and what these models say about opioid addiction policy. I will then examine the history of the opioid epidemic. I will open this by defining opioids, then provide a brief timeline, and then discuss the first American opioid epidemic (which occurred shortly after the Chinese opioid epidemic). Next, I will discuss the factors that led to the current opioid epidemic, such as changes in physician attitudes towards opioids and the introduction of Oxycontin. I will then describe the transition from prescription painkillers to heroin and synthetic opioids, and how they became available across the United States. I will briefly discuss opioid addiction treatment and naloxone, and then give a quick rundown on the state of Ohio. I end the literature review by classifying explanations as either supply side or demand side. In my data and methods chapter, I explain my model, provide summary statistics and visualizations of my data, describe my data sources, and then finally explain my choice of statistical methods. In my results chapter, I run ordinary least squares regressions, and then briefly describe the results. Finally, I provide a conclusion and a copy of some of the code I used to produce this thesis.

```{r censusFunctions}
get_census_table <- function(year, table, survey = "acs1") { # defaults to "acs1" to support legacy code
  df <- get_acs("county",
                year = year,
                table = table,
                survey = survey,
                state = "OH",
                output = "wide",
                cache_table = TRUE) %>%
    rename(county = NAME) %>%
    mutate(county = str_remove(county, " County, Ohio")) %>%
    mutate(year = year,
           survey = survey)
}

# wrapper for get_census_table that handles a year range
get_census_table_multiple_years <- function(years, table, survey = "acs1") { # defaults to "acs1" to support legacy code
  map_df(years, get_census_table, table, survey)
}
```

```{r censusData}
# Pulls down data from census if it is not already stored as a .Rda object in data/
if(!file.exists(here("data", "census.Rda"))){
  
  # DISABILITY
  if(!file.exists(here("data", "disability.Rda"))) {
    
    disability <- bind_rows(get_census_table_multiple_years(2010:2018, "B18101", "acs1"),
                            get_census_table_multiple_years(2012:2018, "B18101", "acs5")) %>%
      mutate(disability_male_percent_under5 = B18101_004E / B18101_003E,
             disability_male_percent_5to17 = B18101_007E / B18101_006E,
             disability_male_percent_18to34 = B18101_010E / B18101_009E,
             disability_male_percent_35to64 = B18101_013E / B18101_012E,
             disability_male_percent_65to74 = B18101_016E / B18101_015E,
             disability_male_percent_75andup = B18101_019E / B18101_018E,
             disability_female_percent_under5 = B18101_023E / B18101_022E,
             disability_female_percent_5to17 = B18101_026E / B18101_025E,
             disability_female_percent_18to34 = B18101_029E / B18101_028E,
             disability_female_percent_35to64 = B18101_032E / B18101_031E,
             disability_female_percent_65to74 = B18101_035E / B18101_034E,
             disability_female_percent_75andup = B18101_038E / B18101_037E) %>%
      select(GEOID, survey, county, year, starts_with("disability"))
    
    save(disability, file = here("data", "disability.Rda"))
    
  } else {
    
    load(here("data", "disability.Rda"))
    
  }
  
  # RACE
  if(!file.exists(here("data", "race.Rda"))) {
    
    race <- bind_rows(get_census_table_multiple_years(2010:2018, "B02001", "acs1"),
                      get_census_table_multiple_years(2010:2018, "B02001", "acs5")) %>%
      mutate(percent_white = B02001_002E / B02001_001E,
             percent_black = B02001_003E / B02001_001E) %>%
      select(GEOID, survey, county, year, percent_white, percent_black)
    
    save(race, file = here("data", "race.Rda"))
    
  } else {
    
    load(here("data", "race.Rda"))
    
  }
  
  # MEDIAN INDIVIDUAL INCOME
  if(!file.exists(here("data", "income_individual.Rda"))) {
    
    income_individual <- bind_rows(get_census_table_multiple_years(2010:2018, "B06011", "acs1"),
                                          get_census_table_multiple_years(2010:2018, "B06011", "acs5")) %>%
      mutate(income_individual_median = B06011_001E) %>%
      select(GEOID, survey, county, year, income_individual_median)
    
    save(income_individual, file = here("data", "income_individual.Rda"))
    
  } else {
    
    load(here("data", "income_individual.Rda"))
    
  }
  
  # MEDIAN HOUSEHOLD INCOME
  if(!file.exists(here("data", "income_household.Rda"))) {
    
    income_household <- bind_rows(get_census_table_multiple_years(2010:2018, "B19013", "acs1"),
                                  get_census_table_multiple_years(2010:2018, "B19013", "acs5")) %>%
      mutate(income_household_median = B19013_001E) %>%
      select(GEOID, survey, county, year, income_household_median)
    
    save(income_household, file = here("data", "income_household.Rda"))
    
  } else {
    
    load(here("data", "income_household.Rda"))
    
  }
  
  # EDUCATION
  if(!file.exists(here("data", "education.Rda"))) {
    
    education <- bind_rows(get_census_table_multiple_years(2010:2018, "B15003", "acs1"),
                           get_census_table_multiple_years(2012:2018, "B15003", "acs5")) %>%
      mutate(education_percent_less_than_highschool = ((B15003_002E + B15003_003E + B15003_004E + B15003_005E + B15003_006E + B15003_007E + B15003_008E + B15003_009E + B15003_010E + B15003_011E + B15003_012E + B15003_013E + B15003_014E + B15003_015E + B15003_016E + B15003_017E + B15003_018E) / B15003_001E),
             education_percent_some_college = ((B15003_019E + B15003_020E + B15003_021E) / B15003_001E),
             education_percent_bachelors_or_more = ((B15003_022E + B15003_023E + B15003_024E + B15003_025E) / B15003_001E)) %>%
      select(GEOID, survey, county, year, starts_with("education"))
    
    save(education, file = here("data", "education.Rda"))
    
  } else {
  
    load(here("data", "education.Rda"))  
  }
  
  census <- left_join(disability, race) %>%
    left_join(education) %>%
    left_join(income_individual) %>%
    left_join(income_household) %>%
    mutate(year = as.numeric(year),
           survey = as_factor(survey),
           county = as_factor(county))
  
  # save the census data as an RDA file
  save(census, file  = here("data", "census.Rda"))
  
} else {
  
  load(here("data", "census.Rda"))
  
}
```

```{r overdoseData}
if(!file.exists(here("data", "overdose.Rda"))) {
  
  # table sourced from https://wonder.cdc.gov/controller/datarequest/D76;jsessionid=C5396F5BBB4351ECF2A36732347C7AFB?stage=results&action=toggle&p=O_show_suppressed&v=true
  # suppressed = 9 or fewer deaths
  # county, year with suppressed rows on
  # all drug-related causes (includes alcohol and the like)
  # fix suppression with https://doi.org/10.2105/AJPH.2014.301900
  # rates per 100,000
  # citation: Centers for Disease Control and Prevention, National Center for Health Statistics. Underlying Cause of Death 1999-2018 on CDC WONDER Online Database, released in 2020. Data are from the Multiple Cause of Death Files, 1999-2018, as compiled from data provided by the 57 vital statistics jurisdictions through the Vital Statistics Cooperative Program. Accessed at http://wonder.cdc.gov/ucd-icd10.html on Apr 5, 2020 10:34:28 AM
  overdose_real <- read_tsv(here("data", "cdc_drug_overdose_deaths.txt")) %>%
    select(-Notes, -`Year Code`) %>% # na in all cases for Notes, Year Code is duplicate of Year
    rename(GEOID = `County Code`,
           year = Year,
           county = County,
           overdose_deaths = Deaths,
           cdc_population = Population,
           fatal_overdose_crude_rate = `Crude Rate`,
           fatal_overdose_age_adjusted_rate = `Age Adjusted Rate`) %>%
    # create dummies for unreliable / suppressed
    mutate(overdose_deaths_suppressed = if_else(overdose_deaths == "Suppressed", TRUE, FALSE),
           fatal_overdose_crude_rate_unreliable = if_else(fatal_overdose_crude_rate == "Unreliable", TRUE, FALSE),
           fatal_overdose_crude_rate_suppressed = if_else(fatal_overdose_crude_rate == "Suppressed", TRUE, FALSE),
           fatal_overdose_age_adjusted_rate_unreliable = if_else(fatal_overdose_age_adjusted_rate == "Unreliable", TRUE, FALSE),
           fatal_overdose_age_adjusted_rate_suppressed = if_else(fatal_overdose_age_adjusted_rate == "Suppressed", TRUE, FALSE)) %>%
    # parse columns with suppressed/unreliable into numbers
    mutate(overdose_deaths = parse_number(overdose_deaths),
           fatal_overdose_crude_rate = parse_number(fatal_overdose_crude_rate),
           fatal_overdose_age_adjusted_rate = parse_number(fatal_overdose_age_adjusted_rate)) %>%
    mutate(county = str_remove(county, " County, OH")) %>%
    # make GEOID a character to make it easy to make it a factor after merging
    mutate(GEOID = as.character(GEOID))
  
  # https://www.cdc.gov/nchs/data-visualization/drug-poisoning-mortality/
  # citation: Rossen LM, Bastian B, Warner M, Khan D, Chong Y. Drug poisoning mortality: United States, 1999–2017. National Center for Health Statistics. 2019.
  overdose_modeled <- read_csv(here("data", "OverdoseDeathRate.csv")) %>%
    filter(State == "Ohio") %>%
    select(GEOID = FIPS,
           county = County,
           year = Year,
           fatal_overdose_modeled_rate = `Model-based Death Rate`,
           cdc_population = Population,
           urban_rural = `Urban/Rural Category`) %>%
    mutate(county = str_remove(county, " County, OH"),
           GEOID = as.character(GEOID),
           year = as.numeric(year),
           urban_rural = as_factor(urban_rural))
  
  overdose <- left_join(overdose_real, overdose_modeled)
  
  save(overdose, file = here("data", "overdose.Rda"))
  
} else {
  
  load(here("data", "overdose.Rda"))
  
}
```

```{r prescribingRate}
# download prescription data for one particular year
prescriptionYear <- function(year) {
  year <- as.character(year)
  df <- glue("https://www.cdc.gov/drugoverdose/maps/rxcounty{year}.html") %>%
    read_html() %>%
    html_node("table") %>%
    html_table() %>%
    mutate(year = year)
  colnames(df) <- c("county", "State", "FIPS", "PrescribingRate", "year")
  df <- df %>%
    filter(State == "OH") %>%
    select(-State, -FIPS) %>%
    mutate(county = str_remove(county, ", OH"))
}

if(!file.exists(here("data", "prescriptions.Rda"))) {
  
  prescriptions <- map_dfr(2006:2018, prescriptionYear) %>%
    mutate(county = as_factor(county),
           PrescribingRate = as.numeric(PrescribingRate),
           year = as.numeric(year))
  
  save(prescriptions, file = here("data", "prescriptions.Rda"))
  
} else {
  
  load(here("data", "prescriptions.Rda"))
  
}
```

```{r grabCountyGeometry, echo=FALSE, message=FALSE, warning=FALSE}
if(!file.exists(here("data", "basemap.Rda"))) {
  
  base_map <- get_acs("county",
                      state = "OH",
                      table = "B02001",
                      geometry = TRUE,
                      output = "wide")
  
  # reduce down to GEOID and geometry
  base_map <- base_map %>%
    select(GEOID, geometry, county = NAME)
  
  save(base_map, file = here("data", "basemap.Rda"))
  
} else {
  
  load(here("data", "basemap.Rda"))
}
```

```{r prepareDF}
if(!file.exists(here("data", "raw_df.Rda"))) {
  
  raw_df <- full_join(overdose, prescriptions) %>% # full join b/c a lot of rows with data only used for making plots / not covered in sample
    left_join(census) %>%
    mutate_at(vars(contains("percent")), ~ .x * 100) %>% # convert percentages to be XX.X instead of .XXX
    mutate(county = as_factor(county),
           GEOID = as_factor(GEOID)) %>%
    ungroup()
  save(raw_df, file = here("data", "raw_df.Rda"))

} else {
  
  load(here("data", "raw_df.Rda"))
  
}
```

```{r mapTheme}
theme_map <- theme_minimal() +
  theme(axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major = element_blank())
```

```{r recodeVariables}
recodeVars <- function(df) {
  df %>%
    rename(County = county,
           Year = year,
           `Overdose Deaths` = overdose_deaths,
           `Population (CDC)` = cdc_population,
           `Fatal Overdose Crude Rate` = fatal_overdose_crude_rate,
           `Fatal Overdose Age-Adjusted Rate` = fatal_overdose_age_adjusted_rate,
           `Are Overdose Deaths Suppressed?` = overdose_deaths_suppressed,
           `Is the Fatal Overdose Crude Rate Unreliable?` = fatal_overdose_crude_rate_unreliable,
           `Is the Fatal Overdose Crude Rate Suppressed?` = fatal_overdose_crude_rate_suppressed,
           `Is the Fatal Overdose Age-Adjusted Rate Unreliable?` = fatal_overdose_age_adjusted_rate_unreliable,
           `Is the Fatal Overdose Age-Adjusted Suppressed?` = fatal_overdose_age_adjusted_rate_suppressed,
           `Fatal Overdose Modeled Rate` = fatal_overdose_modeled_rate,
           `Prescribing Rate` = PrescribingRate,
           `Percent of Males Aged Under 5 with a Disability` = disability_male_percent_under5,
           `Percent of Males Aged 5 to 17 with a Disability` = disability_male_percent_5to17,
           `Percent of Males Aged 18 to 34 with a Disability` = disability_male_percent_18to34,
           `Percent of Males Aged 35 to 64 with a Disability` = disability_male_percent_35to64,
           `Percent of Males Aged 65 to 74 with a Disability` = disability_male_percent_65to74,
           `Percent of Males Aged 75 and Up with a Disability` = disability_male_percent_75andup,
           `Percent of Females Aged Under 5 with a Disability` = disability_female_percent_under5,
           `Percent of Females Aged 5 to 17 with a Disability` = disability_female_percent_5to17,
           `Percent of Females Aged 18 to 34 with a Disability` = disability_female_percent_18to34,
           `Percent of Females Aged 35 to 64 with a Disability` = disability_female_percent_35to64,
           `Percent of Females Aged 65 to 74 with a Disability` = disability_female_percent_65to74,
           `Percent of Females Aged 75 and Up with a Disability` = disability_female_percent_75andup,
           `Percent White` = percent_white,
           `Percent Black` = percent_black,
           `Percent with Highschool or Less` = education_percent_less_than_highschool,
           `Percent with Some College` = education_percent_some_college,
           `Percent with Bachelors or More` = education_percent_bachelors_or_more,
           `Median Individual Income` = income_individual_median,
           `Median Household Income` = income_household_median)
}
```

```{r prepareFinalDF}
df <- raw_df %>%
  # fix urban rural
  filter(!is.na(urban_rural)) %>%
  group_by(county) %>%
  summarize(Density = first(urban_rural)) %>%
  ungroup() %>%
  right_join(raw_df, by = c("county" = "county")) %>%
  recodeVars() %>%
  select(-urban_rural) %>%
  # calculate crude rate when it is not provided
  mutate(`Fatal Overdose Crude Rate` = if_else(!is.na(`Fatal Overdose Crude Rate`),
                                               `Fatal Overdose Crude Rate`,
                                               (`Overdose Deaths` / `Population (CDC)`) * 100000)) %>%
  # create combined fatal overdose rate that includes imputed values
  mutate(`Fatal Overdose Rate (Includes Imputed)` = if_else(is.na(`Fatal Overdose Crude Rate`),
                                                            `Fatal Overdose Modeled Rate`,
                                                            `Fatal Overdose Crude Rate`))
```

```{r prepareDFsForModeling}
df_acs1 <- df %>%
  filter(survey == "acs1")

df_acs5 <- df %>%
  filter(survey == "acs5")
```